---
title: "demux_visualisation"
author: "Luis Montano"
date: '2023-08-21'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: show
    highlight: pygments
    df_print: paged
params:
  newsamplesheet: '~/metadata/10x_annotationsheet_2023-04_ncnb.xlsx'
  datapath: '~/mnt_data/'
  plotpath: '~/mnt_out/scrna/plots/'
  cachedir: '/path/to/cachedir/'
  metadatafolder: '~/metadata/'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Applying demultiplexing to all the libraries

In this script we apply deMULTIplex2 to identify the separate cell lines in each multiplex sample G1-G27. Then we reformat the demultiplexing labels to a standard format which is similar to the cellranger multi output. 

## Dependencies

This script depends on some organiser paths on R/importdata.R and the script R/scutils.R  

## Output

The output of this script is the bash script bash/docker_execute_cellranger_olddata.sh used to run cellranger on the terminal. 

# 1. Load utilities and requirements.

```{r}

source("~/R/scutils.R")

# import data structure with important file paths

# created fulldatapaths list with all the paths to be needed. 
source("./ncnb2_code/R/importdata.R")

cat("Global script parameters:\n")

minfeatures=100
mincells=100
selecttop=5
minfc=0.5
mincpm=3
pcdims=30
vfeatures=2000

```
import external deMULTIplex 2 library

```{r}

devtools::install_github("Gartner-Lab/deMULTIplex2")
```

import datasets
```{r}
fullrecreate=F; fullreload=T
selectednames=names(allcountpaths)[c(1, 3:27)]


solist=lapply(selectednames, function(nm){
  cat("Dataset ", nm, ":", "\nFetching and adapting metadata...\n")
 
  simpleCache(paste_("raw_seurat_", nm), {
  cat("Importing count matrices and HTO assay matrices into Seurat object...\n")
so=CreateSeuratObject(Read10X_h5(allcountpaths[[nm]], use.names = TRUE, unique.features = TRUE)$`Gene Expression`, project=nm, min.cells=mincells, min.features=minfeatures)
print(so)

dmassay=  CreateAssayObject(data= Read10X_h5(allcountpaths[[nm]], use.names = TRUE, unique.features = TRUE)$`Multiplexing Capture`[, so %>% metadata %>% rownames]) 
so[["demux"]]=dmassay

cat("Labeling cell barcodes with dataset name...\n")
so=RenameCells(so, new.names= lapply(so %>% colnames, function(x) strsplit(x, split="-")[[1]][1]) %>% unlist )
so=RenameCells(so, add.cell.id=nm)

so

}, assignToVar="so", recreate=fullrecreate)
  
}) %>% givename(., selectednames)


get.sample.from.bc= Vectorize(function(dataset, bcid){
  (bcrels %>% filter( multiseq_id == bcid, experiment_group == dataset))$sample_name
}, USE.NAMES=F)
  


## Adding basic metadata 
solist=lapply(selectednames, function(nm){
  cat("Dataset ", nm, ":", "\nFetching and adapting demux metadata...\n")
 so=solist[[nm]]
 percent.mt<- PercentageFeatureSet(so, pattern = "^MT-") %>% givecolnames(., nms="percent.mt")
 
 
  simpleCache(paste0("cellmetadata_basic_dataset", nm), {
  #assuming that all the relevant relationships
  md=fread(demuxdatapaths[[nm]]) %>% as.data.frame %>% mutate( dsname=nm,  dsbarcode1=addprefix(Barcode, prefix=paste0(dsname,"_")), dsbarcode=trimone(dsbarcode1), experiment_group=dsname, multiseq_id=Assignment ) %>% col2names(., "dsbarcode")  %>% left_join(., bcrels, by=c("multiseq_id", "experiment_group")) %>% col2names(., "dsbarcode") 

md2=cbind(md , md %>% pull(sample_name) %>% extractname %>% t %>% as.data.frame %>% giverownames(., nms=rownames(md)) ) %>% mutate(replicate=getrep(dsname), stage= stagename(dsname)) 



cat("Incorporating dataset metadata...\n")
fcat(md2 %>% pull(condition) %>% unique)
relevantcells=rownames(md2)
#removing cells that were prefiltered and thus not even subject to demux

md2= cbind(md2,  percent.mt)

md2}, assignToVar="md2", reload=fullreload, recreate=fullrecreate) 
so= so[, rownames(md2)]  
so= AddMetaData(so, metadata=md2)


so
}) %>% givename(., selectednames)
```

Produce a complete table associating, for each experiment, what each barcode name corresponds in terms of demux condition.

the pattern is like this
[samples]
sample_id,cmo_ids
WT_D0_R3,BC54
17q_D0_R3,BC55
17qx_D0_R3,BC59

```{r}

bckey=lapply(selectednames, function(nm){

#### get cellranger config file for the dataset
  
fl= list.files(path=params$metadatafolder, pattern=paste0("cellrangerconfig_", nm, "_"))  


file_content <- readLines(paste0(params$metadatafolder, fl))

# Find the index of the line containing [samples]
start_index <- grep("\\[samples\\]", file_content)

# Extract the lines after [samples]
lines_after_samples <- file_content[(start_index + 1):length(file_content)]

# Create a table using data.table
sample_table <- data.table::fread(text = lines_after_samples, header = TRUE, sep = ",")
sample_table  %>% as.data.frame %>% mutate(dsname=nm)
  
}) %>% Reduce(rbind, .)


bclist=lapply(selectednames, function(nm){

#### get cellranger config file for the dataset
  
fl= list.files(path=params$metadatafolder, pattern=paste0("cellrangerconfig_", nm, "_"))  


file_content <- readLines(paste0(params$metadatafolder, fl))

# Find the index of the line containing [samples]
start_index <- grep("\\[samples\\]", file_content)

# Extract the lines after [samples]
lines_after_samples <- file_content[(start_index + 1):length(file_content)]

# Create a table using data.table
sample_table <- data.table::fread(text = lines_after_samples, header = TRUE, sep = ",") %>% as.data.frame

lii=list()
for(k in 1:nrow(sample_table)){

lii[[sample_table[k, "cmo_ids"]]]=sample_table[k, "sample_id"]
}
lii
  
}) %>% givename(selectednames)


allbarcodes<- bckey %>% pull(cmo_ids)



find.sample= Vectorize(function(ds, bc){

  bc2=gsub("-", "_", bc)
  
  
  if(is.null(bclist[[ds]][[bc2]])){
    
    if(!(bc2 %in% allbarcodes)){
      bc2
    }else{
     bc 
    }
    }else{
     bclist[[ds]][[bc2]]  
    }
}, USE.NAMES=F)
  
  


```



### demux with demMULTIplex2

from https://github.com/Gartner-Lab/deMULTIplex2
demultiplexTags() is the core function of deMULTIplex2. User must provide a tag count matrix where rows are individual cells and columns represent unique sample tags. You can load an example tag matrix from Stoeckius et al. (2018) by calling data(stoeckius_pbmc);tag_mtx <- stoeckius_pbmc. Use ?demultiplexTags in R to check details. It is strongly recommended to pre-filter the matrix to remove majority of the empty droplets for robust classification (See Troubleshooting for details).


```{r}

library(deMULTIplex2) 
simpleCache(paste0("demultiplex2_alldatasets"),{
alldemuxes.raw=lapply(selectednames, function(nm){

res<- demultiplexTags(solist[[nm]]$demux[] %>% as.matrix %>% t, plot.path=params$plotpath, plot.name=paste0("demultiplex2", nm) )

})
names(alldemuxes.raw)= selectednames
alldemuxes.raw
}, assignToVar="alldemuxes.raw", reload=fullrecreate,recreate=fullrecreate)



```

Preparing the demux data to be added to the metadata

```{r}
simpleCache("demultiplex2_all_new_labels", { 
newdemux.full=lapply(selectednames, function(nm){
alldemuxes.raw[[nm]]$final_assign %>% as.data.frame %>% givecolnames(., nms="demux2")
}) %>% givename(selectednames)
newdemux.full
}, assignToVar="newdemux.full", reload=F, recreate=T)


################################################################################
#
################################################################################

simpleCache("demultiplex2_all_processed", { 
newdemux.full.processed=lapply(selectednames, function(nm){

  fcat("formatting demultiplexing info for dataset", nm, "...")
  mat=newdemux.full[[nm]] %>% mutate(demux2.corrected= find.sample(bc=demux2, ds=nm))

demux2conditions=mat %>% pull(demux2.corrected) %>% extractname %>% t %>% as.data.frame 
mat=mat %>% mutate(stage=demux2conditions$stage, cell.line=demux2conditions$cell.line, condition2=demux2conditions$condition, ccondition=demux2conditions$condition, demux2.edited= edit.demux(demux2.corrected), conditiondemux2=get_condition_demux(condition2,demux2.edited ))
}) %>% givename(., selectednames)

}, assignToVar="newdemux.full.processed", reload=F, recreate=T)

```

Incorporate all metadata and all quality control cutoffs and calculations already into metadata. qcglobal indicates whether a cell passed all quality controls

```{r}
########################################################################################
#load previous qc steps
########################################################################################
#load relevant caches from above
fullrecreate=F; fullreload=T
simpleCache(paste0("scdblscores_all_dbr", dbr),assignToVar="alldbscores", reload=T)
simpleCache("cellcycle_information_all_datasets", assignToVar="cc.information", reload=T)
simpleCache("edropletsdata_formatted", assignToVar="alldroplets.ready", reload=T)
simpleCache("demultiplex2_all_new_labels", assignToVar="newdemux.full", reload=T)

solist= lapply(selectednames, function(nm){
  fcat("Dataset", nm)
  #reload basic metadata just in case anything was not loaded yet
  simpleCache(paste0("cellmetadata_basic_dataset", nm), assignToVar="md2", reload=fullreload, recreate=fullrecreate) 
  solist[[nm]] =AddMetaData(solist[[nm]], md2)
  #incorporate cutoffs
  solist[[nm]]= AddMetaData(solist[[nm]], solist[[nm]] %>% metadata %>% mutate( counts.cutoff=qcpars[[nm]]$counts, features.cutoff= qcpars[[nm]]$features, mito.cutoff=qcpars[[nm]]$mito, qcc.pass=((nFeature_RNA>=features.cutoff & nCount_RNA >= counts.cutoff & percent.mt<= mito.cutoff )),scDblFinder.doubletrate=dbr, dbcutoff=getdbcutoff(dsname) ))
  
  
  #incorporate droplets, cc, cutoffs
  
  solist[[nm]] =AddMetaData(solist[[nm]], alldroplets.ready[[nm]] )
  solist[[nm]] =AddMetaData(solist[[nm]], cc.information[[nm]] )
  solist[[nm]] =AddMetaData(solist[[nm]], alldbscores[[nm]] )
  
  #incorporate variable that has the conditions, or failed demux category when present
  solist[[nm]]@meta.data[, "demux_classif"]=solist[[nm]]@meta.data[[demuxvar]] %>% classifydemux
  
  
  solist[[nm]] = AddMetaData(solist[[nm]], solist[[nm]] %>% metadata %>% mutate(stage=stagename(dsname),
                        dayds=paste_(stage, dsname),
                        conditiondemux=get_condition_demux(condition, demux_classif),
                        dbcutoff=getdbcutoff(dsname),
                        edqc.pass= (scdblscore<= dbcutoff & !isemptydroplet),
                        qcpass.global= edqc.pass & qcc.pass))
  
  
################################################################################
#incorporate alternative demux info
################################################################################
 solist[[nm]] = AddMetaData(solist[[nm]], newdemux.full[[nm]]) 
  
}) %>% givename(., selectednames)


simpleCache("allpreqc-metadata", {
allmetadata =solist %>% lapply(., function(x) x %>% metadata %>% mutate(stage=stagename(dsname), replicate=getrep(dsname), demux2.corrected= find.sample(bc=demux2, ds=dsname)) %>% select(-matches("_cnts"))  %>% select(-matches("CMO")) %>% select(-matches("^BC"))   ) %>% Reduce(rbind, .) 
}, assignToVar="allmetadata", recreate=T)
```

Block to edit demux outputs so demux variables are equally formatted

```{r, fig.width=15, fig.height=15}
sw=.5
sca=20

edit.demux= Vectorize(function(x){ 
  if(is.na(x)) {return("Blank")}else{
    if(x=="multiplet") {return("Multiplet")}else{
      if(x=="negative") {return("Unassigned")}else{
        return(x)
      }
  }
  }
}, USE.NAMES=F)




```

Execute the demux formatting  and visualise comparisons between cellranger demux and deMultiplex2

```{r, fig.width=15, fig.height=15}

################################################################################
# formatting demux results
################################################################################

demux2conditions=allmetadata %>% pull(demux2.corrected) %>% extractname %>% t %>% as.data.frame %>% mutate(condition2=condition) %>% select(condition2)
allmetadata.demux2=allmetadata %>% cbind(., demux2conditions) %>% mutate(demux2.edited= edit.demux(demux2.corrected), conditiondemux2=get_condition_demux(condition2,demux2.edited ))


demuxes=allmetadata.demux2 %>% group_by( dsname, conditiondemux, conditiondemux2) %>% summarise( counts=n()) %>% ungroup %>% group_by(dsname) %>% mutate(Freq=counts/sum(counts)) %>% ungroup() %>% as.data.frame 


demuxes2=allmetadata.demux2 %>% names2col(., coln="cell") %>% select(cell, dsname, conditiondemux, conditiondemux2 ) %>% group_by( dsname, conditiondemux, conditiondemux2) %>% summarise( counts=n())  %>% group_by(dsname) %>% mutate(Freq=counts/sum(counts)) %>% ungroup() 

demuxes2=demuxes %>% mutate(id=1:nrow(demuxes2)) %>% pivot_longer(., cols=c("conditiondemux", "conditiondemux2"), names_to="demuxround", values_to="classif") %>% as.data.frame


levs=c( "cWT", "c17q", "c17q1q", "c17q1qMYCN", "Multiplet",  
"Blank","Unassigned")
tpdf(path=params$plotpath, paste0("alluvialv2_demux1_vs_demux2_"), width=pw*5/2*sca/10, height=pw*sca*3/14)
ge=ggplot(demuxes2,
          aes(y = Freq, x=factor(demuxround, levels=c("conditiondemux", "conditiondemux2")), stratum=factor(classif, levels=levs), alluvium=id, fill=factor(classif, levels=levs)))+ 
  geom_flow() +
  geom_stratum(width = sw, alpha=0.5) +
  geom_text_repel(stat = "stratum", aes(label = after_stat(stratum)), min.segment.length=0.1) +
  scale_x_discrete(expand= c(.1, .1), labels = c("cellranger_multi", "demultiplex2"))+
  theme(legend.position="none")+
  facet_wrap(~dsname)+
  theme_classic()
ge
dev.off()

ge
```

