---
title: "Saldana, Montano, et al. -- scRNA-seq: 02_integration of the whole dataset using Seurat"
author: "Luis Montano"
date: '2023-07-18'
output: html_document
params:
  inputdata: '~/path/to/input_data/'
  plotpath: '~/path/to/plots/'
  cachedir: '~/path/to/cachedir/'

 

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Merging all G1-G27 datasets into a full dataset via a simple sparse matrix merge

## Dependencies

This script depends on some organiser paths on R/importdata.R and the script R/scutils.R It also needs the list of post quality control seurat objects from scrna/01_... Rmd as well as the metadata

## Outputs

a seurat merged seurat object comprising the 26 10x datasets, with the corrected metadata (accounting for sample mislabeling). simple_merge_all_filtered_datasets.RData


### Import functions and set parameters. click on the "run all chunks above" icon below  to reload parameters

```{r}
source("~/R/scutils.R")
source("~/R/importdata.R")

nn=30L;
mindist=.3;
md=0.3
res=.5
pcdims=30
vfeatures=2000
selecttop=5
minfc=0.5
mincpm=3
pcdims=30

fdr=0.05
vfeatures.rna=5000
```

## Load list of datasets just after quality control


```{r pressure, echo=FALSE}

simpleCache(paste0("ncnb2-filteredcovariates-seuratobjects_with_markers_vfeatures_", vfeatures), assignToVar="solist2", reload=T)

```


## simple merge of the datasets with each other using the sparse method. 
```{r}

simpleCache("simple_merge_all_filtered_datasets", {
  
  simpleCache("allpostqc-metadata", {
allmetadata.filtered =solist2 %>% lapply(., function(x) x %>% metadata %>% mutate(stage=stagename(dsname), replicate=getrep(dsname)) %>% select(-matches("_cnts"))  %>% select(-matches("CMO")) %>% select(-matches("^BC"))   ) %>% Reduce(rbind, .) 
}, assignToVar="allmetadata.filtered", reload=F, recreate=T)

  
  simpleCache("allpostqc-counts", {
  allcounts= sccore::mergeCountMatrices(
lapply(solist2, function(x) x@assays$RNA@counts
  ))
  }, assignToVar="allcounts", reload=T, recreate=T)  
  
sma=CreateSeuratObject(allcounts, meta.data=allmetadata.filtered, project="all_datasets_merge")


##fixing labeling of 17q cells of the new dataset to be 17q1q based on the finding that 17q1q is the only cell line present

correctcondition=Vectorize(function(ds, cond){
newexpts= seq(14, 27, 1) %>% addprefix(., "G")
if((ds %in% newexpts) && cond=="c17q"){
 "c17q1q" 
}else{
 cond 
}}, USE.NAMES=F)

simpleCache("allpostqc-metadata-correctedcondition",{
sma %>% metadata %>% mutate(ccondition.precorrection=ccondition, ccondition.corrected=correctcondition(dsname, ccondition), ccondition=ccondition.corrected)
}, assignToVar="allmetadata.corrected", reload=T)

sma@meta.data= allmetadata.corrected

sma=AddMetaData(sma, sma %>% metadata %>% mutate(replicate=getrep(dsname)) %>% select(replicate))

}, assignToVar="sma", reload=T)



```


