---
title: "Saldana, Montano, et al. -- scRNA-seq: Dataset integration"
author: "Florian Halbritter"
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: show
    highlight: pygments
    df_print: paged
params:
  inputdata: '~/path/to/input_data/'
  plotpath: '~/path/to/plots/'
  cachedir: '/path/to/cachedir/'
---

# Integration of all datasets past quality control with scvi.
## Dependencies

This script depends on some organiser paths on R/importdata.R and the script R/scutils.R as well as the output simple_merge_all_filtered_datasets.RData produced by scrna/02_1...Rmd

## Outputs

different scvi reductions run on different features, specifically seurat_scviall_RNA_counts_hvg.RData which uses the RNA counts and the highly variable features.  

# Setup

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/rstudio')
```

Load utility scripts, constants, etc. used throughput the project:

```{r}
source("R/setup.R")
source("R/scutils.R")

```
Functions specifically used in this notebook.

```{r}
resultsDir <- function(...) {
  paste0(config$out_root, "/scrna/", ...)
}

runSCVI <- function(so, use_assay = "RNA", use_data = "counts", use_hvg = FALSE, n_epochs = 50, out_dir = resultsDir()) {
  
  soTmp <- so
  if(use_hvg) soTmp <- soTmp[intersect(rownames(soTmp),VariableFeatures(soTmp)), ]
  
  # convert to anndata:
  adata <- convertFormat(soTmp, from="seurat", to="anndata", assay=use_assay, main_layer=use_data, drop_single_values=FALSE)
  print(adata)
  
  # run setup_anndata
  scvi$model$SCVI$setup_anndata(adata, batch_key="replicate")
  
  # create the model
  model <- scvi$model$SCVI(adata)
  
  # train the model
  model$train(max_epochs = as.integer(n_epochs))
  
  # get the latent represenation
  lat <- model$get_latent_representation()
  
  # create DimReduc
  lat <- as.matrix(lat)
  rownames(lat) <- colnames(so)
  
  so[["scvi"]] <- CreateDimReducObject(embeddings = lat, key = "scvi_", assay = use_assay)
  
  return(list(model=model, so=so))  
}

options("RCACHE.DIR"=resultsDir("/rcache/"))
```


# Dataset integration with scVI

## Set up python 

Load the right conda environment:

```{r}
library(reticulate)
use_condaenv("/opt/conda/envs/scvi")
py_config()
```
Load libraries:

```{r}
library(sceasy)
scvi <- import("scvi", convert = FALSE)
```

## Prepare input datasets

Get merged and normalized Seurat object for input data:

```{r}
nHVG <- 5000
cacheSuffix <- paste0("_hvg", nHVG)
if(nHVG == 3000) cacheSuffix <- "" # no suffix for default

simpleCache(paste0("sct_ccvar_all_filtered_datasets",cacheSuffix), {
  simpleCache("simple_merge_all_filtered_datasets", assignToVar="soTmp")
  BiocManager::install("glmGamPoi", update = FALSE, ask = FALSE)
  soTmp <- SCTransform(soTmp, method = "glmGamPoi", variable.features.n=nHVG, vars.to.regress = "ccvar", verbose = TRUE)
  return(soTmp)
}, assignToVar="soMerged", reload=TRUE)
```

Subset full dataset to WT:

```{r}
soWT <- subset(soMerged, ccondition=="cWT")
```


## Run scVI... Full dataset

Execute scVI with the full dataset and plot UMAPs:
(try different combinations of input data and feature selections)

```{r, fig.width=16, fig.height=5}
nDim <- 10
scviInput <- "all"
so <- soMerged
if(scviInput=="wt") so <- soWT
if(scviInput=="test") so <- soWT[sample(nrow(soWT), 1000), sample(ncol(soWT), 1000)]

for(scviHVG in c(TRUE, FALSE)) {
  for(scviAssay in c("RNA")) {  
    for(scviData in c("counts")) { 
      
      n <- paste_(scviInput, scviAssay, scviData, ifelse(scviHVG, "hvg", "all"))
      msg(n)
      
      simpleCache(paste0("seurat_scvi", n), { 
        # train and retrieve latent space:
        scviResult <- runSCVI(so, use_assay = scviAssay, use_data = scviData, use_hvg = scviHVG)
        
        # save scVI model:
        scviResult$model$save(resultsDir("scvi_models/", n), prefix=paste0(n, "_"), save_anndata = FALSE)
      
        return(scviResult$so) 
      }, cacheDir=resultsDir("rcache"), assignToVar="so")

      so <- RunUMAP(so, dims = 1:nDim, reduction = "scvi", n.components = 2)
      p <- DimPlot(so, group.by=c("replicate","ccondition","stage"), label=TRUE) + ggtitle(sprintf("%s (n_dim = %d)", n, nDim))
      
      ggsave(filename = resultsDir("scvi_umap_", n, ".pdf"), plot = p, width=16, height=5)
      print(p)
    }
  }
}
```


# Appendix

**Runtime**

`r time_diff(SETUP_TIME)`

**Session Info**

```{r}
sessionInfo()
```

