---
title: "Saldana, Montano, et al. -- scRNA-seq: Mapping of single cells to the Kameneva and Jansky datasets"
author: "Luis Montano"
date: '2023-08-09'
output: html_document
params:
  inputdata: '~/path/to/input_data/'
  kamenevadata: "~/path/to/kamenevadata/"
  janskydata: "~/path/to/janskydata/"
  plotpath: '~/path/to/plots/'
  cachedir: '~/path/to/cachedir/'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dataset annotation to the Kameneva and Jansky human embryonic adrenal gland datasets

Mapping to the adrenal gland datasets is done in order to assist with dataset annotation, and making sure the divisions of thedataset reflect the biology behind. 

## Dependencies

This script depends on some organiser paths on R/importdata.R and the script R/scutils.R as well as the output simple_merge_all_filtered_datasets.RData produced by scrna/02_1...Rmd

It is also important to download the following datasets and place them inside the inputdata folder:

*Jansky 2021:  Go to https://adrenal.kitz-heidelberg.de/developmental_programs_NB_viz/, select download data and click on "Download Adrenal gland Data" Seurat Object (RDS) to download this.
*Kameneva 2021: http://pklab.med.harvard.edu/artem/adrenal/data/Seurat/adrenal.human.seurat.scrublet.rds


## Outputs

Supplementary figure 2 j, a label transfer alluvial plot for the three datasets, and the metadata with the relevant annotations in kameneva_integration_metadata_fulldataset.RData


```{r}

source("~/R/scutils.R")

# import data structure with important file paths

# created fulldatapaths list with all the paths to be needed. 
source("~/R/importdata.R")

cat("Global script parameters:\n")

minfeatures=100
mincells=100
selecttop=5
minfc=0.5
mincpm=3
pcdims=30
vfeatures=2000
vfeatures.rna=5000
mapping.umapdims=50
```


## Load the full in vitro dataset with markers

```{r}
fullrecreate=F; fullreload=T
simpleCache("simple_merge_all_filtered_datasets", assignToVar="sma", reload=T)

```


## load the Kameneva dataset counts


```{r }

fullrecreate=F; fullreload=T

simpleCache("adameyko_seurat", {
ada= readRDS(file.path(params$kamenevadata, "adrenal.human.seurat.scrublet.rds"))
ada
},assignToVar="ada", recreate=fullrecreate, reload=fullreload)


```

 Processing the kameneva dataset

```{r, echo=FALSE}
fullrecreate=F
fullreload=T
verbose=T
rass="refAssay"
simpleCache("adameyko_seurat_sct_umap", {
  ada=SCTransform(ada, new.assay.name=rass, variable.features.n=vfeatures.rna, assay="RNA" ) 
  DefaultAssay(ada)=rass
  ada=RunPCA(ada, assay=rass)
   ada=RunUMAP(ada, assay=rass, dims=1:mapping.umapdims, verbose=F, return.model=T)
}, recreate=fullrecreate, reload=fullreload, assignToVar="ada")


 ass="sctscvi"    
 ###############################################################################
 # Quantify anchors and already transfer data
 ###############################################################################
   mappedcells=maptoref.pt1(query=sma, refdataset=ada, pdms=1:mapping.umapdims, mapvar="fate2",
                            query.assay=ass,
                            ref.assay=rass,
                            reference.umap="umap",
                            reference.pca="pca",
                              querylabel="invitro", reflabel="kameneva", return.merge.metadata=F, return.query.metadata=TRUE, process.reference=F)
   
   simpleCache(paste_("seurat_adameykomapped", ass, "to", rass), {mappedcells$query}, recreate=fullrecreate, reload=fullreload, assignToVar="sma")
   simpleCache(paste_("anchors_adameykomapped", ass, "to", rass), {mappedcells$anchors}, recreate=fullrecreate, reload=fullreload, assignToVar="anchors")
   
###############################################################################
#perform umap coordinate mapping and export label transfer dataframe   
################################################################################
              
   querymapped=maptoref.pt2(query=mappedcells$query, anchors=mappedcells$anchors, refdataset=mappedcells$ref, pdms=1:mapping.umapdims, mapvar="fate2",
                            query.assay=ass,
                            ref.assay=rass,
                            reference.umap="umap", #which umap onto which we will project cells
                            reference.pca="pca", # which reduction contains the reference's pca, particularly important if many exist
                            reference.reduction="pca", # which reduction to quantify neighbors on
                              querylabel="invitro", reflabel="kameneva", return.merge.metadata=F, return.query.metadata=TRUE, process.reference=F)


sma=AddMetaData(mappedcells$query, metadata=querymapped)

sma@meta.data= sma@meta.data %>% mutate(kamenevatype=mapfun_fate2, type=masktype(mapfun_fate2), type3=masktype3(mapfun_fate2))

 
 simpleCache("seurat_alldatasets_mapped_kameneva", {sma}, assignToVar="sma", reload=F, recreate=T)
 
 simpleCache("kameneva_integration_metadata_fulldataset", {
 sma %>% metadata %>% select(kamenevaumap_1, kamenevaumap_2, mapfun_fate2, mapfun_fate2_predicted.id.score, mapfun_fate2_mapping.score,predicted.id, predicted.id.score, type, kamenevatype)
 }, assignToVar="mappedmeta", recreate=T) 
```


Plot the mapped dataset to verify how the label transfer went


```{r, fig.width=8, fig.height=3}


library(viridis)
umapred="umapog"
    umap.df= umapdf(sma, reduction=umapred)
    rk=paste0(umapred, "_")
    
    umapnn.name=paste0(umapred, "_nn")
umapsnn.name=paste0(umapred, "_snn")
graphchoice=umapsnn.name

    
    sca=10
    rr=600
    ssca=180
    fillvar="clusterfrac"
    clusvarname="seurat_clusters"


polys=clusterhull(NULL, umap.df=umap.df, clusvar=clusvarname, clus=umap.df %>% pull(!!sym((clusvarname))) %>% unique, fillvar="null", linecolor="#000000", size=0.5, reduction.key=rk, fillcols="#FFFFFF00")

pk=ggumap(sma, colorby=clusvarname, labelby=clusvarname, reductions=umapred, reduction.key=rk)+polys+NoAxes()
ps=ggumap(sma, colorby="stage", labelby="stage", reductions=umapred, reduction.key=rk)+ NoLegend()+NoAxes()
pt=ggumap(sma, colorby="type", labelby="type", reductions=umapred, reduction.key=rk)+NoAxes()+squarelegend("type", shape=0, size=6)
pa=ggumap(sma, colorby="adameykotype", labelby="stage", reductions=umapred, reduction.key=rk)+
  NoAxes()+squarelegend("adameykotype", shape=0, size=6)
pm=ggumap(sma, colorby="ccondition", labelby="ccondition", reductions=umapred, reduction.key=rk)+NoAxes()+squarelegend("ccondition", shape=0, size=6)
pt+ps+pm
```

Plot a category transfer alluvial plot to check how each categroy corresponds with cell types

```{r, fig.height=5, fig.width=7}

allcolors[["type"]]= allcolors[["type"]] %>% unlist
g=transferplot(sma, labelvars=c("ccondition", "type"), column.width=0.35, facetvar="stage", colorlist=allcolors)
g
```



 Comparing mappings between jansky and kameneva


```{r}
simpleCache("jansky_processed", {
  janskycounts=readRDS(params$janskydata, "adrenal_gland_Seurat.rds")
janskymeta= readRDS(file.path(params$inputdata, "jansky_adrenal_gland_annot.RDS"))
jansky=CreateSeuratObject(counts=janskycounts, meta.data=janskymeta)
jansky=SCTransform(jansky, method="glmGamPoi", variable.features.n=5000, new.assay.name="refAssay")
jansky=RunPCA(jansky, dims=1:mapping.umapdims)
jansky=RunUMAP(jansky, assay=rass, dims=1:mapping.umapdims, return.model=T)
jansky=AddMetaData(jansky, jansky %>% metadata %>% mutate(janskytype=med_idents))
jansky}, assignToVar="jansky", reload=T)


simpleCache("mapping_metadata_fulldataset_jansky", {
varrjansky="janskytype"
maplist=maptoref.pt1(query=RenameAssays(sma, sctscvi="refAssay"), query.assay="refAssay", refdataset=jansky, pdms=1:dms, mapvar=varrjansky, querylabel="invitro", reflabel="jansky", return.merge.metadata=F, return.query.metadata=TRUE)

  mapcon=maptoref.pt2(query=maplist$query, refdataset=maplist$ref,anchors=maplist$anchors, ref.assay=rass,query.assay=rass, mapvar=varrjansky, querylabel="invitro", reflabel="jansky",  reference.pca="pca", reference.umap="umap", reference.neighbors=umapnn.name.nns, return.merge.metadata=F, return.query.metadata=TRUE, process.reference=F)
  mapcon}, assignToVar="mapcon", reload=T)
  

sma=AddMetaData(sma, metadata=mapcon %>% select(!!sym(mapfun(varrjansky)), !!sym( paste_(mapfun(varrjansky), "predicted.id.score")), !!sym( paste_(mapfun(varrjansky), "mapping.score")) ))

```

Export the metadata from Kameneva 

```{r}

 simpleCache("kameneva_integration_metadata_fulldataset", {
 sma %>% metadata %>% select(kamenevaumap_1, kamenevaumap_2, mapfun_fate2, mapfun_fate2_predicted.id.score, mapfun_fate2_mapping.score,predicted.id, predicted.id.score, kamenevatype )
 }, assignToVar="mappedmeta", recreate=T) 
```


```{r}
janskytypes=jansky %>% metadata %>% pull(janskytype) %>% unique %>% as.character


################################################################################

# Supplementary data figure 2 j alluvial plot comparing mappings between jansky and kameneva

################################################################################

sca=0.5

tp=transferplot(so=NULL, df=sma %>% metadata %>% filter(ccondition=="cWT", !(stage %in% c("D0", "D3"))) %>% mutate(Kameneva.cell.type=kamenevatype, Jansky.cell.type=mapfun_janskytype), labelvars=c("stage", "Jansky.cell.type", "Kameneva.cell.type"), colorlist=allcolors, column.width=0.35)

tpdf(path=params$outpath, "figS2plots_tranferplot_stage_janskytypes_kamenevatypes", wi=pw*sca*1.3, he=pw*sca)
print(tp)
dev.off()



```








