---
title: "Saldana, Montano, et al. -- scRNA-seq: 02_integration of the whole dataset using Seurat"
author: "Luis Montano"
date: '2023-07-18'
output: html_document
params:
  inputdata: '~/path/to/input_data/'
  plotpath: '~/path/to/plots/'
  cachedir: '~/path/to/cachedir/'

 

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Integration of all datasets past quality control with scvi and the adrenal gland label transfer. 
## Dependencies

This script depends on some organiser paths on R/importdata.R and the script R/scutils.R It also requires the output of the script 02_1, 02_2, and 02_3. 

## Outputs

The output of this script is a list of post-qc seurat objects that will be used for integration and other purposes, stored in variable solist2, and inside the RData object "ncnb2-filteredcovariates-seuratobjects_with_markers_vfeatures_vfeatures" where vfeatures is the number of variable features used. 

seurat_alldatasets_scvi_hvg.RData

### Import functions and set parameters. click on the "run all chunks above" icon below  to reload parameters

```{r}
source("~/R/scutils.R")
source("~/R/importdata.R")

nn=30L;
mindist=.3;
md=0.3
res=.5
pcdims=30
vfeatures=2000
selecttop=5
minfc=0.5
mincpm=3
pcdims=30

fdr=0.05
vfeatures.rna=5000
```

## Load the full in vitro dataset with markers

```{r}
fullrecreate=F; fullreload=T

simpleCache("simple_merge_all_filtered_datasets", assignToVar="sma", reload=T)

```


## Importing scvi integration assay, which was run in a GPU server, together with the Kameneva Jansky mappings

```{r}
fullrecreate=T; fullreload=F


simpleCache("seurat_alldatasets_scvi_hvg", {  
simpleCache("seurat_scviall_RNA_counts_hvg", assignToVar="scvi", reload=T)

scvi=RenameAssays(scvi, SCT="sctscvi")
sma@assays$sctscvi=scvi@assays$sctscvi
sma@reductions$scvi=scvi@reductions$scvi

rm(scvi)
gc()


DefaultAssay(sma)="sctscvi"

################################################################################
# Incorporate corrected conditions on metadata
################################################################################

simpleCache("allpostqc-metadata-correctedcondition",{
sma %>% metadata %>% mutate(ccondition.precorrection=ccondition, ccondition.corrected=correctcondition(dsname, ccondition), ccondition=ccondition.corrected)
}, assignToVar="allmetadata.corrected", reload=T)

sma= AddMetaData(sma, metadata= allmetadata.corrected %>% select(ccondition.precorrection, ccondition.corrected, ccondition))
 
################################################################################
# incorporate kameneva mapping metadata
################################################################################
simpleCache("kameneva_integration_metadata_fulldataset", assignToVar="mappedmeta", reload=T)
 
 sma <- AddMetaData(sma, mappedmeta)
  sma@meta.data= sma %>% metadata %>% mutate(type2=type) 
sma
}, assignToVar="sma", reload=fullreload, recreate=fullrecreate)
```

## Session info etc.

Runtime: `r time_diff(SETUP_TIME)`

```{r}
sessionInfo()
```

