---
title: "Saldana, Montano, et al. -- scRNA-seq: 03_WT_marker_analysis"
author: "Luis Montano"
date: '2023-08-07'
output: html_document
params:
  inputdata: '~/path/to/input_data/'
  plotpath: '~/path/to/plots/'
  cachedir: '~/path/to/cachedir/'
  outpath: '~/mnt_out/figures/'
  resourcepath: "~/mnt_resources/"
  fig_width: 6
  fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Main analysis depending exclusively on the WT NC dataset.

## Dependencies

This script depends on some organiser paths on R/importdata.R and the script R/scutils.R It requires to have run all the previous scripts scrna/01..scrna/02 to obtain seurat_alldatasets_scvi_hvg.RData

## Outputs

* The processed WT dataset with clusters seurat_WT_processed_clusters_markers.RData
* cellinfo data structure containing the marker signatures for each cluster  of the WT
* Most of the contents of Fig 1. and some of the contents of Supplementary figure 2. 


 Load utilities
```{r}
source("ncnb2_code/R/scutils.R")
source("ncnb2_code/R/importdata.R")

install.packages("khroma")
library("khroma")
library(RColorBrewer)
```

set up parameters
```{r parame}
vfeatures=5000
pcdims=30
umapdims=10
sz=.02
threed=F #use if a 3d umap is desired
tsca=200

################################################################################
# setting UMAP parameters
################################################################################
  wtumapdims=10
  wtnn=30
  wtag="WT"
  wtmd=0.4
  pval=0.05
  minfc=.3
  fdr=0.05
  selecttop=250
  minrate=0.5
  finalclusvar="WT.clusters"
  
  datte="2023-10-19" # for a new time stamp, run 'chopstring(timestamp(""))' instead
verbose=F
meth="deseq"
groupvar="group1"
pvar="padj"
thresh=pval
fcvar="log_fc"
minrate=.5
replicatecol="replicate"
selecttop=250
################################################################################
#### comparison heatmap parameters
################################################################################  
  numcells=100

################################################################################
# recreation/ reloading controls
################################################################################  
loadcontrols=list()
  loadcontrols$wtmarkers=list(
 loadfulldata="reload",
 processraw="reload",
 calculateclusters="recreate"
)
  
stgnms=c("D0", "D3", "D6", "D9", "D10", "D12", "D14", "D19", "D28")  
allcolors[["stageroma"]]=color("roma")(9) %>% as.character %>% givename(., stgnms)
allcolors[["stageoriginal"]]=c(D0 = "black", D3 = "brown", D6 = "darkslateblue", D9 = "blue", 
D10 = "purple", D12 = "red", D14 = "salmon", D19 = "orange", 
D28 = "yellow")
allcolors[["stagespectral"]]= brewer.pal(n = 11, name = "Spectral")[c(1,2,3,4,5,8,9,10,11)] %>%givename(., stgnms)
allcolors[["stage"]]=allcolors[["stagespectral"]]

```


## Loading Dataset and keeping only WT cells. 


```{r}
fullrecreate=F; fullreload=T


simpleCache("seurat_wt_scvi", {

simpleCache("seurat_alldatasets_scvi_hvg", assignToVar="sma", reload=fullreload, recreate=fullrecreate)

simpleCache("seurat_scviwt_RNA_counts_hvg", assignToVar="scviwt", reload=T)

#filtering all but WT
scwt=sma[, sma %>% metadata %>% filter(ccondition=="cWT") %>% rownames]  
  

scwt
DefaultAssay(scwt)="sctscvi"
}, assignToVar="scwt", recreate=fullrecreate, reload=fullreload)



```

## Declaring several potential UMAP parameter sets

```{r, fig.width=10, fig.height=10}

simpleCache("wtumap_pars", {
wtumap.pars=list(
 one=list(assay="sctharmony", umapname="wtumap.sctharmony", reduction.key=prepare.rk("wtumap.sctharmony"), pcared="sctharmony50dims", nn=wtnn, md=wtmd, umapdims=wtumapdims),
 two=list(assay="sctharwt", umapname="wtumap.sctharwt", reduction.key=prepare.rk("wtumap.sctharwt"), pcared="sctharwt50dims", nn=wtnn, md=wtmd, umapdims=wtumapdims),
 three=list(assay="sctharmony", umapname="wtumap.sctharmony", reduction.key=prepare.rk("wtumap.sctharmony"), pcared="sctharmony50dims", nn=wtnn, md=wtmd, umapdims=20),
 four=list(assay="sctscviwt", umapname="wtumap.scvi5", reduction.key=prepare.rk("wtumap.scvi5"), pcared="scviwt", nn=wtnn, md=wtmd, umapdims=5),
  five=list(assay="sctscviwt", umapname="wtumap.scvi6", reduction.key=prepare.rk("wtumap.scvi6"), pcared="scviwt", nn=wtnn, md=wtmd, umapdims=6),
   six=list(assay="sctscviwt", umapname="wtumap.scvi7", reduction.key=prepare.rk("wtumap.scvi7"), pcared="scviwt", nn=wtnn, md=wtmd, umapdims=7),
    seven=list(assay="sctscviwt", umapname="wtumap.scvi8", reduction.key=prepare.rk("wtumap.scvi8"), pcared="scviwt", nn=wtnn, md=wtmd, umapdims=8),
     eight=list(assay="sctscviwt", umapname="wtumap.scvi9", reduction.key=prepare.rk("wtumap.scvi9"), pcared="scviwt", nn=wtnn, md=wtmd, umapdims=9),
      nine=list(assay="sctscviwt", umapname="wtumap.scvi10", reduction.key=prepare.rk("wtumap.scvi9"), pcared="scviwt", nn=wtnn, md=wtmd, umapdims=9),
 ten=list(assay="sctscvi", umapname="wtumap.fullscvi5", reduction.key=prepare.rk("wtumap.fullscvi5"), pcared="scvi", nn=wtnn, md=wtmd, umapdims=5),
 eleven=list(assay="sctscvi", umapname="wtumap.fullscvi6", reduction.key=prepare.rk("wtumap.fullscvi6"), pcared="scvi", nn=wtnn, md=wtmd, umapdims=6),
 twelve=list(assay="sctscvi", umapname="wtumap.fullscvi7", reduction.key=prepare.rk("wtumap.fullscvi7"), pcared="scvi", nn=wtnn, md=wtmd, umapdims=7),
 thirteen=list(assay="sctscvi", umapname="wtumap.fullscvi8", reduction.key=prepare.rk("wtumap.fullscvi8"), pcared="scvi", nn=wtnn, md=wtmd, umapdims=8),
 fourteen=list(assay="sctscvi", umapname="wtumap.fullscvi9", reduction.key=prepare.rk("wtumap.fullscvi9"), pcared="scvi", nn=wtnn, md=wtmd, umapdims=9),
       fifteen=list(assay="sctscvi", umapname="wtumap.fullscvi1015", reduction.key=prepare.rk("wtumap.fullscvi1015"), pcared="scvi", nn=wtnn, md=wtmd, umapdims=10),
      sixteen=list(assay="sctscvi", umapname="wtumap.fullscvi1016", reduction.key=prepare.rk("wtumap.fullscvi1016"), pcared="scvi", nn=40, md=wtmd, umapdims=10),
       seventeen=list(assay="sctscvi", umapname="wtumap.fullscvi1017", reduction.key=prepare.rk("wtumap.fullscvi1017"), pcared="scvi", nn=50, md=wtmd, umapdims=10),
        eighteen=list(assay="sctscvi", umapname="wtumap.fullscvi1018", reduction.key=prepare.rk("wtumap.fullscvi1018"), pcared="scvi", nn=40, md=.5, umapdims=10),         nineteen=list(assay="sctscvi", umapname="wtumap.fullscvi1019", reduction.key=prepare.rk("wtumap.fullscvi1019"), pcared="scvi", nn=40, md=.6, umapdims=10), twenty=list(assay="sctscvi", umapname="wtumap.fullscvi1020", reduction.key=prepare.rk("wtumap.fullscvi1020"), pcared="scvi", nn=50, md=.5, umapdims=10),
 twentyone=list(assay="sctscvi", umapname="wtumap.fullscvi921", reduction.key=prepare.rk("wtumap.fullscvi921"), pcared="scvi", nn=50, md=.5, umapdims=9), 
 
 ####### parameter set 22 used in final wt projection
 
  twentytwo=list(assay="sctscvi", umapname="wtumap.fullscvi822", reduction.key=prepare.rk("wtumap.fullscvi822"), pcared="scvi", nn=50, md=.5, umapdims=8, umapclusres=0.08)
)
wtumap.pars}, assignToVar="wtumap.pars", recreate=F, reload=T)

```


## set up parameters to preferred parameter set and calculating umap

```{r}
for(umapchoice in 22){ 
  
  ur=wtumap.pars[[umapchoice]]$umapname
  ass=wtumap.pars[[umapchoice]]$assay
  red=wtumap.pars[[umapchoice]]$pcared
  rk=wtumap.pars[[umapchoice]]$reduction.key
  nn=wtumap.pars[[umapchoice]]$nn
  md=wtumap.pars[[umapchoice]]$md
  umapdims=wtumap.pars[[umapchoice]]$umapdims
  
    wtur=wtumap.pars[[umapchoice]]$umapname
  wtass=wtumap.pars[[umapchoice]]$assay
  wtred=wtumap.pars[[umapchoice]]$pcared
  wtrk=wtumap.pars[[umapchoice]]$reduction.key
  wtnn=wtumap.pars[[umapchoice]]$nn
  wtmd=wtumap.pars[[umapchoice]]$md
  wtumapdims=wtumap.pars[[umapchoice]]$umapdims
  if(!is.null(wtumap.pars[[umapchoice]]$umapclusres)){
  wtumapclusres=wtumap.pars[[umapchoice]]$umapclusres
  }
          scwt=RunUMAP(scwt, reduction=wtred, assay=wtass, dims=1:wtumapdims, n.neighbors=wtnn, min.dist=wtmd, reduction.name=wtur, reduction.key=wtrk, return.model=T)
}
```


```{r}

sca=10
rr=600
varr="stage"
varrlab=varr
wtumapchoice=22

wtur=wtumap.pars[[wtumapchoice]]$umapname
  wtass=wtumap.pars[[wtumapchoice]]$assay
  wtred=wtumap.pars[[wtumapchoice]]$pcared
  wtrk=wtumap.pars[[wtumapchoice]]$reduction.key
  wtnn=wtumap.pars[[wtumapchoice]]$nn
  wtmd=wtumap.pars[[wtumapchoice]]$md
  wtumapdims=wtumap.pars[[wtumapchoice]]$umapdims
tpng(path=params$plotpath, paste_("UMAPwt", wtur, varr, "umap_par_set", umapchoice), wi=w*sca, he=w*sca, res=rr)
tit=paste(wtur, wtass, wtred, "nn", wtnn, "md", wtmd, "dims", umapdims)
wp=ggumap(scwt, colorby=varr, labelby=varrlab, reduction=wtur, reduction.key=wtrk, colorlist=allcolors, ssca=350)+ggtitle(tit)+NoAxes()+NoLegend()

dev.off()
wp
```

### plot final UMAP choice. 

```{r, fig.width=10, fig.height=10}
tit=paste0("hvgparset ", wtumapchoice," ", wtred," " ,wtumapdims, " dimensions", "dims", wtumapdims, "md", wtmd, "nn", wtnn)

sca=1.5

tpdf(path=params$plotpath, paste_("wtUMAP_alldatasets_allgenes_assay", wtass, "dims", wtumapdims, "md", wtmd, "nn", wtnn), wi=pw*sca*2.5, he=pw*sca*2.5)

scwt=AddMetaData(scwt, metadata= sma[, colnames(scwt)] %>% metadata %>% select( type3))

pl4=lapply(c("stage", "WT.clusters", "type3", "replicate", "Phase", "kamenevatype"#, 
             ,"mapfun_janskytype",
             "type3"
             ), function(varr){
  ggumap(scwt, colorby=varr, labelby=varr, sz=sz,ssca=300,  reduction=wtur, reduction.key=wtrk, colorlist=allcolors , color.labels.by=varr)+
 ggtitle(paste(wtur,"by",varr, tit))+NoAxes()
}) %>% Reduce('+', .)
print(pl4)
dev.off()

pl4

  ggumap(scwt, colorby="mapfun_janskytype", labelby="mapfun_janskytype", sz=.1,ssca=60,  reduction=wtur, reduction.key=wtrk, legend.size=10, color.labels.by="mapfun_janskytype")

```


### Make quality control plots. Supplementary figure 2 a-g supplementary figure 3 c-h
```{r, fig.width=12, echo=F}

sca=3
tpdf(path=params$outpath, "fig0plots_qcplots_UMAP_WT_by_dataset", wi=pw*sca*1.7, he=pw*sca*1.2)
print(qcplots(scwt, plotred=wtur, colorby = "orig.ident"))
dev.off()

sca=3
tpdf(path=params$outpath, "fig0plots_qcplots_UMAP_fulldata_by_dataset", wi=pw*sca*1.5, he=pw*sca*1.2)
print(qcplots(sma, plotred=ur, colorby = "orig.ident"))
dev.off()
```

### add module scores for the signatures of different cell types reported in Kameneva et al, 2021

```{r, fig.width=12, fig.height=12}


nolivermarkers=dropfield(adamarkers, "liver")

scwt=AddModuleScore(scwt, features=nolivermarkers, name=nolivermarkers %>% names)
#Warning: The following features are not present in the object: EFFL7, not searching for symbol synonyms
#Warning: The following features are not present in the object: CYP17A1, CYP21A2, not searching for symbol synonyms
#Warning: The following features are not present in the object: HBA2, not searching for symbol synonyms
#Warning: The following features are not present in the object: HNF4A, ITIH1, not searching for symbol synonyms


```

### Figure 1 c plotting marker strength of different cell types across the WT umap


```{r, fig.width=8, fig.height=8}

viridischoice="D"

markercols=adamarkers[c("stemcells", "sympathoblasts", "mesenchyme", "SCP", "chromaffin")] %>% c(., "FN1") %>% Reduce(c, .)

plm=lapply(markercols, function(m){
  p=FeaturePlot(scwt, features=m, reduction=wtur, raster=T)+scale_color_viridis(option=viridischoice)+NoAxes()
  p
}) %>% Reduce('+', .)+plot_layout(ncol=6, nrow=7)

sca=3
tpdf(path=params$outpath, "UMAPwt_celltypemarkers_featureplots_fulldataset", wi=pw*sca, he=pw*sca)
plm
dev.off()
plm
```


### Figure 1 e,f blend plots of mes, sym and scp module score signatures

```{r, fig.width=12, fig.height=12}


sca=2
bt=0.3
blendcols= c("lightgrey", "#ff0000", "#0000ff")
blendcols2= c("lightgrey", "#ff0000", "#00cc00")
blendcols3=c("lightgrey", "#00cc00", "#0000ff")

blendcols.messym=c("lightgrey", "#0000ff", "#00cc00")
blendcols.scpsym=c("lightgrey", "#ff0000", "#00cc00")
blendcols.messcp=c("lightgrey", "#0000ff", "#ff0000")


fpss=FeaturePlot(scwt, features=c("mesenchyme3", "sympathoblasts2"), reduction=wtur,  blend=T,blend.th=bt, col=blendcols.messym, raster=T)+NoAxes()
fpsm=FeaturePlot(scwt, features=c( "SCP1" , "sympathoblasts2"), reduction=wtur,  blend=T,blend.th=bt, col=blendcols.scpsym, raster=T)+NoAxes()
fpscm=FeaturePlot(scwt, features=c("mesenchyme3", "SCP1"), reduction=wtur,  blend=T, blend.th=bt,col=blendcols.messcp, raster=T)+NoAxes()
tpdf(path=params$outpath, "FeatureblendWT_scp_sympathoblasts", wi=pw*sca*2, he=pw*sca*1.5)#, res=rr); 
fpss/fpscm/fpsm; dev.off()

fpss/fpscm/fpsm



################################################################################
# another colorscheme
################################################################################
basecol="#444444"
blendcolsms= c(basecol, allcolors[["type3"]]["mesenchyme"],  allcolors[["type3"]]["sympathoblasts"])
blendcolsmsc= c(basecol, allcolors[["type3"]]["mesenchyme"] ,allcolors[["type3"]]["SCP"] )
blendcolsss=c(basecol, allcolors[["type3"]]["SCP"], allcolors[["type3"]]["sympathoblasts"])
fpssn=FeaturePlot(scwt, features=c("mesenchyme3", "sympathoblasts2"), reduction=wtur,  blend=T,blend.th=bt, col=blendcolsms, raster=T)+NoAxes()
fpsmn=FeaturePlot(scwt, features=c( "SCP1" , "sympathoblasts2"), reduction=wtur,  blend=T,blend.th=bt, col=blendcolsss, raster=T)+NoAxes()
fpscmn=FeaturePlot(scwt, features=c("mesenchyme3", "SCP1"), reduction=wtur,  blend=T, blend.th=bt,col=blendcolsmsc, raster=T)+NoAxes()
tpdf(path=params$outpath, "FeatureblendWT_scp_sympathoblasts_dark_contrast", wi=pw*sca*2, he=pw*sca*1.5)#, res=rr); 
fpssn/fpscmn/fpsmn; dev.off() 

```

## defining clustering parameters and performing clustering on umap; making glassworkplots

```{r}

knn=wtnn
wtumapclusres=.4
umapred=wtur #declared in previous blocks
umapnn.name=paste0(umapred, "_nngraph")
umapsnn.name=paste0(umapred, "_snngraph")
graphchoice=umapnn.name
clusvarname=paste0(graphchoice, "_res.", as.character(wtumapclusres))

scwt=FindNeighbors(scwt, reduction=umapred, dims=c(1,2), graph.name=c(umapnn.name,umapsnn.name), k.param=knn)
scwt=FindClusters(scwt, res=wtumapclusres, graph.name=graphchoice, algorithm=4)# leiden



varr=clusvarname #right after the clustering also seurat_clusters works fine"
library(viridis)
    umapscwt= umapdf(scwt, reduction=wtur)
    varrlab=varr
    sca=10
    rr=600
    ssca=180
    fillvar="clusterfrac"

   colorpoints=varr
## generating multiple polygons one for each cluster
polys=clusterhull3(NULL, umap.df=umapscwt, clusvar=clusvarname, clus=umapscwt %>% pull(!!sym((clusvarname))) %>% unique, fillvar="null", linecolor="#000000", size=0.2, reduction.key=wtrk, fillcols="#FFFFFF00", concavity=2)

pk=ggumap(scwt, colorby="seurat_clusters", reductions=umapred, reduction.key=wtrk)+ polys+ NoLegend()+NoAxes()
ps=ggumap(scwt, colorby="stage", reductions=umapred, reduction.key=wtrk)+ polys+ NoLegend()+NoAxes()
pp=ggumap(scwt, colorby="type2", reductions=umapred, reduction.key=wtrk)+ polys+ NoLegend()+NoAxes()
pka=ggumap(scwt, colorby="kamenevatype", reductions=umapred, reduction.key=wtrk)+ polys+ NoLegend()+NoAxes()
pk+ps+pp+pka

```



## Fusing clusters based on markers and label transfer, and arranging clusters in roughly temporal order . Figure 1 b, d


```{r}


################################################################################
#  rearranging clusters according to approximate developmental order or another 
# arbitrary 
################################################################################
stage.ordered.clusters=scwt %>% metadata %>% mutate(stagenumber=get.stage.number(stage)) %>% group_by(!!sym(clusvarname) ) %>% summarise(stagemean=mean(stagenumber)) %>% arrange(stagemean) %>% pull(!!sym(clusvarname)) %>% as.vector %>% as.numeric

###relabel, reorder and merge clusters *roughly* from day 0 to day 19

arrclusters=triagecells.multi(scwt, c(2,3,17,23), c(7,24), c(1,8,6), 15,10,21, 18,16,13,20, c(14, 5, 9),12, c(22, 19),c(11,4)) %>% addprefix(., "C")
cluslabel=paste0(clusvarname, ".arranged")
scwt@meta.data[, cluslabel]=arrclusters 
scwt@meta.data[,finalclusvar]= arrclusters
numclusters_arranged= arrclusters %>% unique
allcolors[[cluslabel]]=namenums(randomcolors(numclusters_arranged %>% length), prefix="C", zero=F)
allcolors[[finalclusvar]]=namenums(randomcolors(numclusters_arranged %>% length), prefix="C", zero=F)
umapscwt=scwt %>% umapdf(., reductions=wtur)
polys2=clusterhull3(NULL, umap.df=umapscwt, clusvar=cluslabel, clus=umapscwt %>% pull(!!sym((cluslabel))) %>% unique, fillvar="null", linecolor="#000000", size=0.3, reduction.key=wtrk, fillcols="#FFFFFF00")
polys4=clusterhull3(NULL, umap.df=umapscwt, clusvar=cluslabel, clus=umapscwt %>% pull(!!sym((cluslabel))) %>% unique, fillvar="WT.clusters", linecolor="#000000", size=0.3, reduction.key=wtrk, fillcols=allcolors[["WT.clusters"]])


pa=ggumap(scwt, colorby=finalclusvar, labelby=finalclusvar,  colorlist=allcolors, reductions=wtur, reduction.key=wtrk)+ polys+ NoLegend()+NoAxes()

ps=ggumap(scwt, colorby="type3", labelby=finalclusvar,  colorlist=allcolors, reductions=umapred, reduction.key=wtrk)+ polys2+ NoLegend()+NoAxes()

pww=ggumap(scwt, colorby="type3", labelby=finalclusvar,  colorlist=allcolors, reductions=umapred, reduction.key=wtrk)+ polys4+ NoLegend()+NoAxes()
pa
ps
pw


lapply(c("stage", "type3"), function(cvarr){
sca=2
tpdf(path=params$outpath, paste0("UMAPwt_clusters_coloredby_",cvarr),wi=pw*sca, he=pw*sca)#, res=rr)

ps=ggumap(scwt, colorby=cvarr, labelby="WT.clusters",  colorlist=allcolors, reductions=umapred, reduction.key=wtrk)+ polys2+ NoLegend()+NoAxes()
print(ps)

dev.off()
})


```

### Calculate markers for the clusters using the DElegate::FindAllMArkers2 function

```{r}

options(future.globals.maxSize = 2 * 10^9)
fullreload=F; fullrecreate=T


simpleCache(paste_("seurat_WT_with_delegatemarkers",wtur, "clusterres", umapclusres, "method", meth, datte), {
fcat(paste0("finding markers using the DElegate package...\n") ) 

simpleCache(paste_("cbmarkers_WT",wtur, "clusterres", wtumapclusres,"on_clusters_of", cluslabel,"method", meth, "minrate",minrate, datte), {
  
 m= DElegate::FindAllMarkers2(object=scwt, group_column=finalclusvar, replicate_column=replicatecol, method=meth, min_fc=1, min_rate=minrate)
 m

}, assignToVar="m", reload=T, recreate=F)
clusters=allcolors[[finalclusvar]] %>% names


 ## arrange and filter best markers
    ltx=lapply(clusters, function(x) m %>% filter(!!sym(groupvar)==x,!!sym(pvar)<=thresh, !!sym(fcvar)>=minfc) %>% arrange(-!!sym(fcvar)))
    
  if(is.null(selecttop)){
  lt=lapply(clusters, function(x) m %>% filter(!!sym(groupvar)==x,!!sym(pvar)<=thresh, !!sym(fcvar)>=minfc) %>% arrange(-!!sym(fcvar)))
  topstr=""  
}else{
  lt=lapply(clusters, function(x) m %>% filter(!!sym(groupvar)==x,!!sym(pvar)<=thresh, !!sym(fcvar)>=minfc) %>% arrange(-!!sym(fcvar)) %>% head(n=selecttop))
}
    ms=ltx %>% Reduce(rbind, .)
    mp=lt %>% Reduce(rbind,.)
    

cat("Number of filtered markers: ", nrow(mp), "(", (nrow(mp)/nrow(m))*100, "% markers post filtering)\n")

scwt[["RNA"]]@misc$markers <- m
scwt[["RNA"]]@misc$top_markers <- mp
scwt[["RNA"]]@misc$significant_markers <- ms
scwt[["RNA"]]@misc$marker.params=list(minfc=minfc, pvalue=pval, minrate=minrate, method=meth, group.var.name=groupvar, significance.var.name=pvar, foldchange.var.name=fcvar, grouping.col=finalclusvar, cluster.res= umapclusres, replicate_column=replicatecol, date=datte, num.top.markers=selecttop, min.rate=minrate)
scwt <- GetResidual(scwt, features = scwt$RNA@misc$top_markers$feature, verbose = F)
fwrite(get.significant.markers(scwt), file=file.path(params$outpath, timestamp("marker_table_WT.csv" )))


  scwt
}, assignToVar="scwt", reload=fullreload, recreate=fullrecreate)

```


### store wt dataset with curated clusters and markers calculated

```{r, fig.width=10, fig.height=10}

simpleCache("seurat_wt_alldatasets_scvi_hvg_fullyprocessed", {scwt}, assignToVar="scwt", reload=T, recreate=F)

```

## Cell type marker analysis
### Heatmap of cluster markers
#### Preparing heatmap components


```{r}
fullrecreate=F
fullreload=T
ncells.show=300
smeth="seurat"
groupvar="group1"
fcvar="log_fc"
minrate=0.5
minfc=1
selecttop=250


################################################################################
#prepare cells for heatmap: order cells, prepare markers and gap information, store everything inside cellinfo object
################################################################################

simpleCache(paste0("wtheatmap_clusterinfo_genes_lfct_", minfc, "_numtop_", selecttop, "minrate", minrate, "_", datte, "seriationmethod", smeth, "ncells", ncells.show ), {

wtcellinfo= seriatecells(scwt, clusvar=finalclusvar.wt, markers=NULL, clusters=allcolors[[finalclusvar.wt]] %>% names, meth=smeth,  groupvarname=groupvar,lfcvarname=fcvar, extended.output=T, deduped=T, ncells=ncells.show)
wtcellinfo}, assignToVar="wtcellinfo", recreate=fullrecreate, reload=fullreload)


# selected list of known developmental neural crest markers to search for in the actual list of markers

mrkrs=c("NANOG", "POU5F1", "FGF17","EGR1","BPI","KLF4", "TBXT", "MOXD1","HAND1", "ALK", "NEFM",
        "CNTNAP2","TFAP2B", "POSTN", "PLP1", "SOX10", "CRABP1",
         "DLK1",
        "NEGR1",
        "ANK2",
        "NRCAM",
        "GAP43",
        "ERBB3", "PHOX2B", "ONECUT1", "ASCL1", "VIM",
         "ELAVL4", "MSX2", "COL1A1", "TWIST1", "FN1", "GATA3", "TH", "DBH", "SIX1")


pha=cellinfo.heatmap(scwt, refresh.cellinfo(wtcellinfo) %>%  add.cell.annotation(scwt, ., "stage"), assay="sctscvi", genes.to.label=c( adjust.markers(wtcellinfo, 2)$markers), color.labels.by = "WT.clusters", colorlist=allcolors, name="   ")

sca=15
nm=paste0("pheatmap_WT.clusters_highestlabeled")
tpng(path=params$outpath, nm, width=w*sca*3/2, height=w*sca/10*11, res=rr)
print(pha)
dev.off()

phb=cellinfo.heatmap(scwt, refresh.cellinfo(wtcellinfo) %>%  add.cell.annotation(scwt, ., "stage"), assay="sctscvi", genes.to.label=mrkrs, colorlist=allcolors, name="   ")

################################################################################
# Figure 1 C wt cluster markers heatmap
################################################################################

sca=15
nm=paste0("fig1plots_pheatmap_WT.clusters_highestlabeled")
tpng(path=params$outpath, nm, width=w*sca*3/2, height=w*sca/10*11, res=rr)
print(pha)
dev.off()

sca=1
nm=paste0("fig1plots_pheatmap_WT.clusters_selectedmarkers")
tpdf(path=params$outpath, nm, width=pw*sca*3/2, height=pw*sca/10*11)#, res=rr)
print(phb)
dev.off()

pha
```

### supplementary fig 2 h point plot of hox genes

```{r}

  clus2table=tmp %>% filter(cluster==2) %>% slice_max(avg_log2FC, n=10000)
  hoxnumbers2=clus2table %>% pull(gene) %>% grepl("HOX",.) %>% which
  clus2hox=clus2table[hoxnumbers2,] %>% as.data.frame
  
  
  clus3table=tmp %>% filter(cluster==3) %>% slice_max(avg_log2FC, n=10000)
  hoxnumbers3=clus3table %>% pull(gene) %>% grepl("HOX",.) %>% which
  clus3hox=clus3table[hoxnumbers3,] %>% as.data.frame
  
  allhox= rownames(sma)[sma %>% rownames  %>% grepl("^HOX",.)]


prephox= function(x) if(grepl('-', x)) return(substr(x, 4,8)) else( paste0(substr(x, 1,3), substr(x, 5,5), substr(x, 4,4)))
alphahox= sapply(allhox, prephox) %>% sort %>% names
boxhox=allhox %>% sort

boxes=c("A","B","C","D")
purehox= boxhox[!grepl("-AS", boxhox)]
purehoxletters= substr(purehox, 4,4)
purehoxnumbers= substr(purehox, 5,7)
purehox.orderednumbers=lapply(boxes, function(box){
  purehoxnumbers[purehoxletters==box] %>% as.numeric %>% sort
  
}) %>% Reduce(c, .)


purehox.clusteredbynumbers=lapply(1:11, function(b){
  purehox[purehoxnumbers==b]
  
}) %>% Reduce(c, .)

get.letter=Vectorize(function(x) substr(x, 4,4), USE.NAMES=F)
get.number=Vectorize(function(x) substr(x, 5,6) %>% as.numeric, USE.NAMES=F)

finalhox=paste("HOX", purehoxletters, purehox.orderednumbers, sep="")

hoxtab= join_meta_exp(scwt, genes=finalhox, assay=wtass)

tidyhox= hoxtab %>% pivot_longer(., cols=all_of(finalhox), names_to="gene", values_to="expression") %>% mutate(box=get.letter(gene), hoxnumber=get.number(gene)) 

groupedhox=tidyhox %>% mutate(is.expressed= expression>0) %>% group_by(gene, WT.clusters) %>% summarise(meanexp=mean(expression), pct.expressed=sum(is.expressed)/n())

sca=1
tpdf(path=params$outpath, paste0("bubbleplot_hox"), width=pw*sca*.8, height=pw*sca*1.3)

hoxplot=ggplot(groupedhox)+geom_point(aes(x=factor(WT.clusters, levels=allcolors[["WT.clusters"]] %>% names), y=factor(gene, levels=purehox.clusteredbynumbers %>% rev), size=pct.expressed, color=meanexp))+theme_classic()+scale_color_viridis()+rotatex(90)+scale_size(range=c(0.2,4))

print(hoxplot)
dev.off()


#hs= c()
lapply(1:4, function(x){
b=boxes[x]
totnumbers=(hoxtab2  %>% filter(box==b) %>% pull(hoxnumber) %>% unique %>% length)
sca=totnumbers
fontbase=20
tpng(path=params$plotpath, paste0("bubbleplot_hox"), width=w*4*4, height=w*sca, res=600)
print(hoxplot4(hoxtab2 %>% mutate(hoxnumber= hoxnumber+.05*totnumbers, position=sapply(hoxnumber, function(x) apposition[x], USE.NAMES=F)) %>% filter(box==b) ))
dev.off()        
})


```

### Marker expression feature maps

```{r, fig.width=15, fig.height=15}

##########################################################################
#   Figure 1 d.  Marker expression feature umaps
##########################################################################

reverseflag=T
#first check if reverseflag exists. if not, set it to false. then  run this.
if(reverseflag==FALSE){
scwt@reductions$umap@cell.embeddings[, "UMAP_1"]= -scwt@reductions$umap@cell.embeddings[, "UMAP_1"]
  reverseflag=TRUE
}

cmappink=c("deeppink4", "yellow")

cmapchoice=cmappink
viridischoice="D"
markerselectionplot=FeaturePlot(scwt, c(mrkrs),col=cmapchoice, combine=FALSE, reduction=ur)

allmarkerplots= lapply(markerselectionplot, function(x) x+NoAxes()+scale_color_viridis(option=viridischoice)) %>% Reduce('+', .)
sca=50

rr=800
tpng(path=params$plotpath, paste_("UMAP", ur, "plot_wt_marker_selection_plots_tighter_viridis_", viridischoice), he=sca*w, wi=sca*w, res=rr)
allmarkerplots[[11]]
dev.off()
allmarkerplots
```


### Storing the wild type object


```{r}
simpleCache("seurat_WT_processed_clusters_markers", {scwt}, 
            #recreate=T,
            reload=T,
            assignToVar="scwt")


```


### Supplementary Data Figure 2 i. heatmap of cell types and their markers in both datasets

```{r}
###############################################################################

# Supplementary Data Figure 2 i. heatmap of cell types and their markers in both datasets

###############################################################################
fullreload=T
fullrecreate=0
#step 1. mapping the in vitro cells to the Kameneva 2021 dataset


#step 2. save/ load calculated mappings
## step 2.a return a cache that contains the precalculated mappings

#dataset that contains both reference and query

simpleCache("adameyko_seurat_sct_umap", recreate=fullrecreate, reload=fullreload, assignToVar="ada")

ass.kameneva="refAssay"
ass.ncnb="sctscvi"


#######################################
########NCNB dataset mapping markers
#######################################
##for each cell type, retrieve the  top mapping cellsin the in vitro dataset

plotcats=adamarkers %>% names
mapvar=mapfun("fate2")
scorevar=mapvar %>% paste0(., "_predicted.id.score")
scorevar.map=mapvar %>% paste0(., "_mapping.score")
adamarkersvector.named= adamarkers %>% unlist 
adamarkersvector= adamarkers %>% unlist %>% unname                           
#list with the cells with top scores for each cell type in plotcats

ncnblist=lapply(plotcats, function(tt){
  #get the cell names
  metadata(scwt) %>% filter(kamenevatype==tt) %>% arrange(-!!sym(scorevar)) %>% head(n=numcells) %>% select(!!sym(scorevar), !!sym(scorevar.map)) ##slice max is not doing the job of retrieving only numcells.
  }) %>% givename(., plotcats)

ncnbcellist=lapply(ncnblist, function(l){l %>% rownames}) %>% givename(., plotcats)
ncnbscorelist=lapply(ncnblist, function(l){l %>% pull(!!sym(scorevar))}) %>% givename(., plotcats)
ncnbmaplist=lapply(ncnblist, function(l){l %>% pull(!!sym(scorevar.map))}) %>% givename(., plotcats)  


ncfullmat0=as.matrix(scwt[adamarkersvector,]@assays[[ass.ncnb]]@data)
missing.genes=setdiff(adamarkersvector, ncfullmat0 %>% rownames ) ##incorporate rows for missing markers in the data

ncfullmat=ncfullmat0 %>% rbind(. ,matrix(nrow=length(missing.genes), ncol= ncol(ncfullmat0), data=0) %>% giverownames(., nms=missing.genes) %>% givecolnames(., nms=colnames(ncfullmat0)))

ncgenemeans=apply(ncfullmat, 1, mean)
ncgenesd=apply(ncfullmat, 1, sd)

##retrieving data matrices for each cell group
ncmats=lapply(1:length(ncnbcellist), function(x){
  fcat(names(ncnbcellist[x]))
  cl=ncnbcellist[[x]]
  if(length(cl)>0){
  ncfullmat[adamarkersvector, cl]
  }else{return(NULL)}
}) %>% givename(.,plotcats)
ncmatsc=lapply(1:length(ncnbcellist), function(x){
  
  
  if(!is.null(ncmats)){
  (ncmats[[x]]-ncgenemeans)/ncgenesd
  }else{return(NULL)}
}) %>% givename(., plotcats)


#######################################
########ADAMEYKO dataset cell type example cells
#######################################
# sample a number of cells of each type (potentially can be improved by ranking cells by their weight in the signature but the sampling is for representation)

nostem.markers=dropfield(adamarkers, "stemcells")
nostemcats= nostem.markers %>% names

adafullmat0=as.matrix(ada[adamarkersvector,]@assays[[ass.kameneva]]@data)
missing.genes=setdiff(adamarkersvector, adafullmat0 %>% rownames ) ##incorporate rows for missing markers in the data

adafullmat=adafullmat0 %>% rbind(. ,matrix(nrow=length(missing.genes), ncol= ncol(adafullmat0), data=0) %>% giverownames(., nms=missing.genes) %>% givecolnames(., nms=colnames(adafullmat0)))
rm(adafullmat0)

adavar=demapfun(mapvar)
adacellist=lapply(nostemcats, function(tt){
  #get the cell names
  fcat(tt)
  #sampling cells 
  metadata(ada) %>% filter(!!sym(adavar)==tt) %>% slice_sample(n=numcells) %>% rownames
  }); names(adacellist)=nostemcats
##retrieving data matrices for these cells. 
adagenemeans=apply(adafullmat, 1, mean)
adagenesd=apply(adafullmat, 1, sd)

adamats=lapply(adacellist, function(cl){
  
  if(length(cl)>0){
  adafullmat[adamarkersvector, cl]
  }else{return(NULL)}
}) %>% givename(., nostemcats)


adamatsc=lapply(adacellist, function(cl){
  
  if(length(cl)>0){
  (adafullmat[adamarkersvector, cl]-adagenemeans)/adagenesd
  }else{return(NULL)}
}) %>% givename(., nostemcats)


######################################################
# concatenate cells of the same type from ncnb and adameyko datasets and create split and legend variables on the way
#####################################################
colgroups=list()
coltypes= list()
colgroupsc=list()
coltypesc= list()
ordmat=lapply(plotcats, function(ct){

  colgroups[[ct]] <<-list(nc=colnames(ncmats[[ct]]), ada=colnames(adamats[[ct]]))
  coltypes[[ct]] <<-list(nc=rep(ct, length(colnames(ncmats[[ct]]))), ada=rep(ct, length(colnames(adamats[[ct]]))))
  cat(cbind(adamats[[ct]], ncmats[[ct]] ) %>% dim, "\n")
  cbind(adamats[[ct]], ncmats[[ct]] )

})

ordmat2=lapply(plotcats, function(ct){
  #for each category, splice together matrices from

  colgroupsc[[ct]] <<-list(nc=colnames(ncmatsc[[ct]]), ada=colnames(adamatsc[[ct]]))
  coltypesc[[ct]] <<-list(nc=rep(ct, length(colnames(ncmatsc[[ct]]))), ada=rep(ct, length(colnames(adamatsc[[ct]]))))
  cat(cbind(adamatsc[[ct]], ncmatsc[[ct]] ) %>% dim, "\n")
  cbind(adamatsc[[ct]], ncmatsc[[ct]] )

})

ordmat=ordmat2
##################################################
# prepare the annotation for the columns
#################################################
# create corresponding dataset labels
colanno=colgroups %>% lapply(., function(x) rbind(cbind(x$ada, rep("Kameneva", length(x$ada))), cbind(x$nc, rep("in_vitro", length(x$nc))))) 
#getting rid of stemcells category which is not present in the dataset of adameyko and thus not mapped. 
colanno= colanno[c(1:6, 8:12)] %>% Reduce(rbind, .)
colnames(colanno)= c("cells","origin")
rownames(colanno)=colanno[, "cells"]
colanno=as.data.frame(colanno)
colanno[["cell_type"]]= Reduce(c,lapply(coltypes, function(x)  c(x$ada, x$nc)))

##calculate gaps for each type and dataset
colgaps1=colgroups %>% lapply(., function(x) c(length(x$ada), length(x$nc))) %>% Reduce(c, .) %>% cumsum
#calculate gaps for each type only
colgaps2=colgroups %>% lapply(., function(x) sum(length(x$nc), length(x$ada))) %>% Reduce(c, .) %>% cumsum


################################################################################
# prepare the annotation for the row markers
################################################################################
#types to which each marker belongs
adamarkertypes= lapply(1:length(adamarkers), function(x){rep(names(adamarkers[x]), length(adamarkers[[x]]))}) %>% Reduce(c, .)
markeranno= cbind(adamarkersvector.named %>% unname, adamarkertypes) %>% as.data.frame %>% col2names(., "V1") %>% mutate(cell_type=factor(adamarkertypes, levels= names(adamarkers))) %>% select(cell_type) 

rowgaps=markeranno %>% pull(cell_type) %>% factor %>% as.integer %>% diff %>% as.logical %>% which

###############################
# PLOTTING matrix
###############################
cmap= c('#0B9988','#f7f7f7','#F9AD03')
  br=seq(-3,3,.1)
colls=colorRampPalette(cmap, space="rgb")(length(br))
sca=9;
rr=600;
row_gaps=rowgaps #set to NULL if unwanted
 

allcolors$origin=c(Kameneva="#AB00AB", in_vitro="#F99F00")
allcolors$cell_type=c(HSC_and_immune = "#bebada", intermediate_mesoderm = "#CE9B70", 
sympathoblasts = "#fdb462", endothelium = "#B64BD4", cortex = "#87b1d2", 
HSC_and_immune = "firebrick", kidney = "#E6F288FF", mesenchyme = "#8dd3c7", 
chromaffin = "#438900", SCP = "#FDA6D8FF", melanocytes = "#999999", 
liver = "#FF7BBAFF", stemcells="firebrick", erythroid="red")

m=Reduce(cbind, ordmat[c(1:6,  8:12)])  %>% apply(., c(1,2), as.numeric)

```
### Draw the parallel comparison heatmap
```{r, fig.width=7, fig.height=5 }

sca=1
tpdf(path=params$outpath, "heatmap_markercomparison_onlycelltypegaps_normrow_sctdata_gaps_type_dataset", width=pw*3*sca, height=pw*2.2*sca)
ph=ComplexHeatmap::pheatmap(m[, rownames(colanno)],
                            #scale="row",
                            breaks=br,
                            gaps_col=colgaps1,
                            gaps_row=row_gaps,
                            cluster_row=FALSE, 
                            cluster_col=FALSE,
                            show_colnames=F,
                            border_color=NA,
                            col=colls,
                            use_raster=T,
                            annotation_row=markeranno,
                            annotation_col=colanno %>% select(cell_type, origin),
                            annotation_colors=allcolors[c("cell_type", "origin")], 
                            fontsize=10
                            ) 
print(ph)
dev.off()


```



figure data


```{r}


fwrite(umapdf(scwt, red=wtur) %>% select(wtumapfullscvi822_1,
wtumapfullscvi822_2,WT.clusters, stage, kamenevatype,type3,SCP1,sympathoblasts2
,mesenchyme3), file=file.path(params$outpath, "figuredata_wt_metadata_w_umap.csv"), row.names=T   )


fwrite(wtcellinfo$cell.annotation %>% rbind(., wtcellinfo$marker.annotation), file=file.path(params$outpath, "figuredata_wt_heatmap_annotation.csv"), row.names=T   )

```

## Session info etc.

Runtime: `r time_diff(SETUP_TIME)`

```{r}
sessionInfo()
```
