---
title: 'Saldana, Montano, et al. -- ATAC-seq: Unsupervised analysis' 
author: 'Florian Halbritter'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: show
    highlight: pygments
    df_print: paged
---

# Intro

In this notebook, we perform a simple principal component analysis of the global chromatin accessibility in all samples and conditions (*Fig. 8c* of the paper).


# Setup

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/rstudio')
```

Load utility scripts, constants, etc. used throughput the project:

```{r, include=FALSE}
source("R/setup.R")
source("R/atacutils.R")

library(pheatmap)
library(RColorBrewer)
```

Load required input data from previous steps:

```{r load_caches}
dA <- latestCache(CACHE_ATAC_META)
dds <- latestCache(CACHE_ATAC_DDS)
```

Normalize count data and remove constant rows:

```{r norm_data}
dVst <- SummarizedExperiment::assay(DESeq2::vst(dds))
dVst <- dVst[rowSums(dVst!=0)>0 & apply(dVst,1,sd)>0,] # remove constant rows
dim(dVst)
```


# Principal component analysis

Run PCA and format results as table combined with sample annotation:

```{r calc_pca}
pcaRes <- prcomp(t(dVst), scale.=T, center=T)
pData <- as.data.table(pcaRes$x, keep.rownames="library_name") 
pData <- merge(pData, dA, by="library_name")
print(head(pData))
```

Calculate arrow positions between geometric mean of each timepoint / condition:

```{r calc_arrow_pos}
arrowData <- rblapply(1:(length(DAY_ORDER)-1), function(i) {
	s <- DAY_ORDER[i]
	e <- DAY_ORDER[i+1]
	
	arrowData <- merge(pData[day_fac==s, .(srcPC1=median(PC1), srcPC2=median(PC2)), by=.(condition,treatment)], pData[day_fac==e, .(dstPC1=median(PC1), dstPC2=median(PC2)), by=.(condition,treatment)], by=c("condition","treatment"))

	data.table(s=s, e=e, arrowData, actual=TRUE)
}, "start")
print(head(arrowData))
```

Add mock connections for the 17q1qMYCN prior to DOX treatment (= identical to 17q1q):

```{r extend_arrow_data}
arrowData <- rbind(arrowData, arrowData[condition=="17q1q" & start<3,.(start, s, e, condition="17q1qMYCN", treatment="dox", srcPC1, srcPC2, dstPC1, dstPC2, actual=FALSE)])
arrowData[condition=="17q1qMYCN" & e=="D9", `:=`(c("dstPC1","dstPC2"),arrowData[condition=="17q1qMYCN" & s=="D9", .(srcPC1,srcPC2)])]
```

Three different ways of laying out the PCA plots:

```{r plot_pca}
p <- ggplot(pData, aes(x=PC1, y=PC2, color=day_fac)) + ylab(sprintf("Principal component 2 (%.1f%%)", summary(pcaRes)$importance[2,2]*100)) + xlab(sprintf("Principal component 1 (%.1f%%)", summary(pcaRes)$importance[2,1]*100)) + defTheme(noLegendTitle=T) + geom_point(size=2) + scale_color_manual(values=COLOR_PALETTES$day) + theme(panel.border=element_rect(colour="black", fill=NA))
print(p)

p <- ggplot(arrowData, aes(x=srcPC1,y=srcPC2,xend=dstPC1,yend=dstPC2,linetype=actual, color=condition)) + ylab(sprintf("Principal component 2 (%.1f%%)", summary(pcaRes)$importance[2,2]*100)) + xlab(sprintf("Principal component 1 (%.1f%%)", summary(pcaRes)$importance[2,1]*100)) + defTheme(noLegendTitle=T) + scale_color_manual(values=COLOR_PALETTES$condition) + geom_segment(data=arrowData, arrow = arrow(type="closed", length = unit(0.03, "npc"))) + scale_linetype_manual(values=c("TRUE"="solid","FALSE"="dashed")) + theme(panel.border=element_rect(colour="black", fill=NA))
print(p)

p <- ggplot(pData, aes(x=PC1, y=PC2, color=condition)) + ylab(sprintf("Principal component 2 (%.1f%%)", summary(pcaRes)$importance[2,2]*100)) + xlab(sprintf("Principal component 1 (%.1f%%)", summary(pcaRes)$importance[2,1]*100)) + defTheme(noLegendTitle=T)  + scale_color_manual(values=COLOR_PALETTES$condition) + themeBoxed(axis.text=element_blank(), axis.ticks=element_blank(), panel.grid.major=element_line(size=0.25))
p <- p + geom_point(aes(shape=day_fac),color="black") 
p <- p + geom_segment(aes(x=srcPC1,y=srcPC2,xend=dstPC1,yend=dstPC2), data=arrowData, alpha=0.75, linejoin="mitre", arrow = arrow(type="closed", length = unit(0.085, "npc"))) + scale_linetype_manual(values=c("TRUE"="solid","FALSE"="dashed"))
p <- p + facet_wrap(~factor(condition,levels=CONDITION_ORDER), ncol=2)
print(p)
```
```{r}
ggsave(filename = figuresDir("figure_atac_pca.pdf"), plot = p, width = 6, height = 4, dpi = 600, unit = "in")
fwrite(pData[,.(condition, day_fac, PC1, PC2)], file = figuresDir("figure_atac_pca_data.csv"))
```




# Appendix

**Runtime**

`r time_diff(SETUP_TIME)`

**Session Info**

```{r}
sessionInfo()
```
