---
title: 'Saldana, Montano, et al. -- ATAC-seq: Differential accessibility analysis and modules' 
author: 'Florian Halbritter'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: show
    highlight: pygments
    df_print: paged
---


# Intro

In this notebook, we perform differential accessibility analyses between different developmental stages and different conditions (cell lines) and then group the identified differentially accessibile regions into "chromatin modules".
These analyses directly feed into Fig 9a and Supplementary Figs. 12a-c of the paper.


# Setup

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/rstudio')
```

Load utility scripts, constants, etc. used throughput the project:

```{r, include=FALSE}
source("R/setup.R")
source("R/atacutils.R")

library(DESeq2)
```

Functions / constants specifically used in this notebook.

```{r def_func}
prepHeatmapData <- function(d, do_scale=TRUE, do_cap=abscap) {
	hmData <- d
	dn <- dimnames(hmData)	
	if(do_scale==TRUE) hmData <- t(apply(hmData, 1, scale))
	dimnames(hmData) <- dn
	do_cap(hmData)
}

plotHMorder <- function(d, row_cols, n, row_lbls, dist_meth="euclidean", clust_meth="complete", do_cap=abscap, do_scale=TRUE, rc=NULL, show_colnames=FALSE, annot_cols=c("condition", "day_fac"), row_annot=peaksDt, cellwidth=8, col_annot=dA,...) {		
	hmData <- prepHeatmapData(d, do_scale=do_scale, do_cap=do_cap)
	
	hmCol <- c("white", "#EEEEEE", "black") 
	if(do_scale) hmCol <- rev(RColorBrewer::brewer.pal(9, "RdBu"))
	
	addCuts <- F
	if(is.null(rc)) {
		rc <- fastcluster::hclust(dist(hmData, method=dist_meth), clust_meth)
		addCuts <- T
	}

	o <- col_annot[colnames(hmData),][order(factor(condition, levels=CONDITION_ORDER), day_fac, batch), library_name]
	hmData <- hmData[,o]
	
	colAnnot <- dtToDf(col_annot[colnames(hmData),unique(c("library_name",annot_cols)),with=F]) #, batch, lane
	rowAnnot <- dtToDf(row_annot[rownames(hmData),intersect(c("rid", "module_id", row_cols),colnames(peaksDt)),with=F])
				
	if(addCuts) {
		for(k in c(3:12)) {
			x <- cutree(rc, k)
			rowAnnot[, paste_("cut",k)] <- paste0("xyz",x)
		}
	}
				
	p <- ComplexHeatmap::pheatmap(hmData, 
					col=hmCol, 
					cellwidth=cellwidth,
					gaps_col=col_annot[colnames(hmData),cumsum(table(factor(condition,levels=CONDITION_ORDER)))], 
					border_color=NA, 
					row_labels=row_lbls, 
					show_colnames=show_colnames, show_rownames=T, 
					use_raster=T, raster_resize_mat=F, raster_device="CairoPNG", raster_by_magick=F, 
					main=n, 
					cluster_col=FALSE, cluster_row=rc, 
					annotation_col=colAnnot, 
					annotation_row=rowAnnot, 
					annotation_colors=COLOR_PALETTES, 
					...)
	
	p
}

ATAC_P_THRESH <- 0.005
ATAC_LFC_THRESH <- log2(1.5) 
ATAC_DIST_METH <- "euclidean" 
ATAC_CLUST_METH <- "ward.D2"
BG_NAME <- "BG"
```

Load required input data from previous steps:

```{r load_caches}
peaksDt <- latestCache(CACHE_ATAC_PEAKS)
dA <- latestCache(CACHE_ATAC_META)
dds <- latestCache(CACHE_ATAC_DDS)
d <- SummarizedExperiment::assay(DESeq2::vst(dds))
```



# Differential aceessibility

## Run differential tests

Use the pre-defined DESeq2 models to perform pair-wise comparisons between different conditions (cell lines) at the same day, or between developmental stages of the same conditon.

```{r run_differential_tests}
darRes <- versionedCache(CACHE_ATAC_DDS_RES, instruction={
	# extract the DESeq2 results for a given comparison:
	runDiffTest <- function(dds, cont) {
		if(length(cont)!=3) return(NULL)
		res <- as.data.table(DESeq2::results(dds, 
		                                     contrast=cont, 
		                                     pAdjustMethod="BH", 
		                                     lfcThreshold=lfcThresh, 
		                                     independentFiltering=FALSE, 
		                                     format="DataFrame", 
		                                     alpha=pThresh, 
		                                     tidy=TRUE))
		return(res)
	}	
	
	grpVar <- "sample_group"
	res <- data.table()
	
	# compile results for two types of comparisons:

	# (i) condition X vs Y, stratified by timepoint		
	tmp <- rblapply(dayOrder, function(strat) { 	
			rblapply(conditionOrder, function(bg) {
				bgGrp <- annot[day_fac==strat & condition==bg, unique(get(grpVar))]		
				rblapply(setdiff(conditionOrder, bg), function(fg) { 
					fgGrp <- annot[day_fac==strat & condition==fg, unique(get(grpVar))]
					if(length(fgGrp)==0 | length(bgGrp)==0) return(NULL) # note: no +dox at D0 & D3						
					msgF("\tstratum %s:\t%s vs %s", strat, paste0(fgGrp, collapse="+"), paste0(bgGrp, collapse="+"))
					cont <- c(grpVar, fgGrp, bgGrp)	
					res <- runDiffTest(dds, cont)
					if(!is.null(res)) cbind(res, type="conds")
				}, "fg")
			}, "bg")
		}, "stratum", cores=4)
	res <- rbind(tmp, res, use.names=T)
	
	# (ii) all pairwise, stratified by condition:		
	tmp <- rblapply(conditionOrder, function(strat) {		
			rblapply(dayOrder, function(fgI) { 
				rblapply(setdiff(dayOrder, fgI), function(bgI) { 
					bg <- annot[day_fac==bgI & condition==strat, unique(get(grpVar))]			
					fg <- annot[day_fac==fgI & condition==strat, unique(get(grpVar))]
					msgF("stratum %s:\t%s (%s) vs %s (%s)", strat, fgI, paste0(fg, collapse="+"), bgI, paste0(bg, collapse="+"))
					cont <- c(grpVar, fg, bg)			
					res <- runDiffTest(dds, cont)
					if(!is.null(res)) cbind(res, type="pairwise")
				}, "bg")				
			}, "fg")
		}, "stratum", cores=4)
	res <- rbind(tmp, res, use.names=T)
		
	res <- unique(res)		
	
	# some cosmetics:
	res[, `rid` := row]
	res[, `row` := NULL] 	
	res[, `dir` := sign(log2FoldChange)]
	res[, `cmp` := sprintf("%s|%s_vs_%s",stratum,fg,bg)]

  # define significant DARs:
  res[, sig:= is.finite(padj) & padj<=pThresh & abs(log2FoldChange)>=lfcThresh]

	return(res)
}, buildEnvir=list(
  dds=dds,
  dayOrder=DAY_ORDER,
  conditionOrder=CONDITION_ORDER,
  pThresh=ATAC_P_THRESH,
  lfcThresh=ATAC_LFC_THRESH,
  annot=dA[, .(day_fac, condition, library_name, sample_group)]
))

print(darRes[,.(N=sum(sig), perc=sum(sig)/.N),by=.(stratum, fg, bg)])
print(darRes[sig==T,.(N=length(unique(rid)))][, .(n_sig=N, n_ns=nrow(peaksDt)-N)])
```

Add DAR information to main peak annotation table:

```{r extend_dar_info}
tmp <- dcast(darRes[sig==T,], rid~cmp, value.var="sig", fun.aggregate=function(x) !is.null(x) & length(x)>0)
darCols <- colnames(tmp)[-1]
forEach(darCols, function(n) suppressWarnings(peaksDt[, paste(n):=NULL]))
peaksDt <- merge(peaksDt, tmp[,c("rid",darCols), with=F], by="rid", all.x=T)
forEach(darCols, function(n) peaksDt[, paste(n) := !is.na(get(n)) & get(n)])
peaksDt[, sig:=rowSums(.SD[,darCols,with=FALSE], na.rm=T)>0] # significant in at least one comparison?
rm(tmp)
```



## Define region sets of interest

We define set of peaks for consideration in enrichment analyses and plots annotations, e.g., for all DARs in a particular condition.

```{r def_rois_1}
regionSetsOfInterest <- c()
mrg <- merge(peaksDt, darRes[sig==T,], by="rid")
```


Add ROIs for all up-regulated DARs per condition (across all stages):

```{r def_rois_2}
tmp <- sapply(darCols, function(darCol) {
	dtToGr(mrg[cmp==darCol & dir>0,], metaCols=c("rid"))
})
names(tmp) <- paste_(names(tmp), "up")
regionSetsOfInterest[names(tmp)] <- tmp
```

Add ROIs for all down-regulated DARs per condition (across all stages):

```{r def_rois_3}
tmp <- sapply(darCols, function(darCol) {
	dtToGr(mrg[cmp==darCol & dir<0,], metaCols=c("rid"))
})
names(tmp) <- paste_(names(tmp), "down")
regionSetsOfInterest[names(tmp)] <- tmp
```

Add ROIs for all DARs per condition:

```{r def_rois_4}
tmp <- sapply(2:length(CONDITION_ORDER), function(condI) {
	fg <- CONDITION_ORDER[condI]
	bg <- CONDITION_ORDER[1] # or vs prev condition?  
	selCmps <- grep(paste_(fg,"vs", bg,"(up|down)"), names(regionSetsOfInterest), value=T)
	unique(unlist(GenomicRanges::GRangesList(regionSetsOfInterest[selCmps]), use.names=F))
}, simplify=FALSE)
names(tmp) <- CONDITION_ORDER[-1]
regionSetsOfInterest[names(tmp)] <- tmp
peaksDt <- merge(peaksDt, dcast(rblapply(tmp, function(x) data.table(rid=x$rid)), rid~id), by="rid", all.x=TRUE)
forEach(names(tmp), function(x) {
	peaksDt[, (x):=!is.na(get(x))]
})
```

Up DARS compared to WT:

```{r def_rois_5}
for(curDir in c("up","down")) {
	tmp <- sapply(2:length(CONDITION_ORDER), function(condI) {
		fg <- CONDITION_ORDER[condI]
		bg <- CONDITION_ORDER[1] # vs WT
		selCmps <- grep(paste_(fg,"vs",bg,curDir), names(regionSetsOfInterest), value=T)
		unique(unlist(GenomicRanges::GRangesList(regionSetsOfInterest[selCmps]), use.names=F))
	}, simplify=FALSE)
	names(tmp) <- paste_(CONDITION_ORDER[-1],curDir)
	regionSetsOfInterest[names(tmp)] <- tmp
	suppressWarnings(forEach(names(tmp), function(x) peaksDt[, (x):=NULL]))
	peaksDt <- merge(peaksDt, dcast(rblapply(tmp, function(x) data.table(rid=x$rid)), rid~id), by="rid", all.x=TRUE)
	forEach(names(tmp), function(x) {
		peaksDt[, (x):=!is.na(get(x))]
	})
}
```
Add "background" ROI, i.e. all peaks:

```{r def_rois_6}
regionUniv <- dtToGr(peaksDt, metaCols=c("rid"))
names(regionUniv) <- peaksDt[,rid]
regionSetsOfInterest[[BG_NAME]] <- regionUniv
```



## Overview of DAR results

Export tables:

```{r export_tabs}
print(darRes[sig==T,.N,by=.(cmp,dir)])
fwrite(darRes[sig==T,.N,by=.(type,fg,bg,cmp,dir)], file=resultsDir("dar_numbers.csv"))
```

Summarize number of DARs for different comparisons:

```{r prep_dar_overview}
hmData <- sapply(CONDITION_ORDER, function(cond) {
	dtToDf(dcast(darRes[stratum==cond & type=="pairwise",.(sig=any(sig)),by=.(rid,bg,fg)][,.(N=sum(sig)), by=.(bg,fg)], bg~fg, value.var="N"))
}, simplify=F)
names(hmData) <- paste_("pairwise",names(hmData))
hmData <- c(hmData, list(
	conds_vs_wt = dtToDf(dcast(darRes[sig==TRUE & type=="conds" & bg=="WT",.(sig=any(sig)),by=.(rid,stratum,fg)][,.(N=sum(sig)), by=.(stratum,fg)], fg~stratum, value.var="N")),
	conds_vs_17q = dtToDf(dcast(darRes[sig==TRUE & type=="conds" & bg=="17q",.(sig=any(sig)),by=.(rid,stratum,fg)][,.(N=sum(sig)), by=.(stratum,fg)], fg~stratum, value.var="N")),
	conds_vs_17q1q = dtToDf(dcast(darRes[sig==TRUE & type=="conds" & bg=="17q1q",.(sig=any(sig)),by=.(rid,stratum,fg)][,.(N=sum(sig)), by=.(stratum,fg)], fg~stratum, value.var="N"))
))
```

Make heatmaps:

```{r heatmap_dar_overview, fig.width=12}
pList <- NULL
xcols <- function(d, rows, cols) {	
	for(n in setdiff(cols,colnames(d))) d[,n] <- rep(NA,nrow(d))
	for(n in setdiff(rows,rownames(d))) d[n,] <- rep(NA,ncol(d))	
	return(as.matrix(d[rows,cols]))
}
forEach(grep("pairwise",names(hmData),value=T), function(n) {
	hmDataX <- xcols(hmData[[n]], DAY_ORDER, DAY_ORDER)	
	hmDataS <- apply(hmDataX, 2, prettyIntegers)
	hmDataS[hmDataS=="NA"] <- ""
	p <- ComplexHeatmap::pheatmap(log10(hmDataX+1), scale="none", breaks=seq(0,ceiling(log10(max(unlist(hmData), na.rm=T)+1)),length.out=21), col=viridis::magma(9), cellwidth=24, cellheight=12, border_color="white", main=n, cluster_col=F, cluster_row=F, display_numbers=hmDataS, number_format="%s", number_color="black")
	pList <<- pList + p
})
# plot heatmaps next to each other:
ComplexHeatmap::draw(pList)
```

```{r heatmap_dar_overview_2, fig.width=8}
pList <- NULL
forEach(grep("pairwise",names(hmData),invert=T, value=T), function(n) {
	hmDataX <- xcols(hmData[[n]], CONDITION_ORDER, DAY_ORDER)	
	hmDataS <- apply(hmDataX, 2, prettyIntegers)
	hmDataS[hmDataS=="NA"] <- ""
	p <- ComplexHeatmap::pheatmap(log10(hmDataX+1), scale="none", breaks=seq(0,ceiling(log10(max(unlist(hmData), na.rm=T)+1)),length.out=21), col=viridis::magma(9), cellwidth=24, cellheight=12, border_color="white", main=n, cluster_col=F, cluster_row=F, display_numbers=hmDataS, number_format="%s", number_color="black")
	pList <<- pList + p
})
# plot heatmaps next to each other:
ComplexHeatmap::draw(pList)
```


## Definition of chromatin modules

### 1. Region modules R1-R6 = differential during WT differentiation

```{r atac_modules_1}
# use all samples for module definition:
moduleSamples <- dA[, library_name]
# ... and all regions found significant in at least one wild-type comparison:
moduleRegs <- darRes[type=="pairwise" & sig==T, unique(rid)] 
K <- 6

tmp <- versionedCache("atac_modules", instruction={		

	hmData <- prepHeatmapData(input_data[module_regs, module_samples], do_scale=TRUE)

	# * calculate distance matrix and perform clustering:
	distMat <- dist(hmData, method=dist_meth)
	rc <- fastcluster::hclust(distMat, clust_meth)
	
	return(list(hm_data=hmData, row_clust=rc))
	
}, buildEnvir=list(
  input_data=d, module_regs=moduleRegs, module_samples=moduleSamples, dist_meth=ATAC_DIST_METH, clust_meth=ATAC_CLUST_METH
))
```

Assign each DAR to one of those "modules":

```{r atac_modules_2}
peaksDt[, module_id_tmp:=paste0("tmp",cutree(tmp$row_clust, k=K)[rid])]
setkey(peaksDt, rid)
```

Figure out an order of region modules that puts the modules in roughly temporal order:

```{r atac_modules_3}
darOrderTmp <- rownames(tmp$hm_data)[tmp$row_clust$order]
pData <- prepHeatmapData(d[darOrderTmp,], do_scale=TRUE)
pData <- melt(as.data.table(pData, keep.rownames="rid"), variable.name="library_name", measure.vars=colnames(pData))
pData[, module_id_tmp:=peaksDt[rid, module_id_tmp]]
pData <- merge(pData, dA, by="library_name")

tmp <- dtToDf(dcast(pData[condition=="WT", mean(value), by=.(day_fac, module_id_tmp)], module_id_tmp~day_fac, value.var="V1"))[,DAY_ORDER]

modOrderTmp <- (rownames(tmp)[seriation::get_order(seriation::seriate(tmp))])
round(tmp[modOrderTmp,])
```

```{r atac_modules_4}
darOrder <- unlist(lapply(split.data.frame(d[darOrderTmp,], f=peaksDt[darOrderTmp, factor(module_id_tmp, levels=modOrderTmp)]), function(mat) {
	rownames(mat)[seriation::get_order(seriation::seriate(mat))]
}), use.names=F)
modOrder <- paste0("R",1:K)
moduleIds <- structure(modOrder, names=modOrderTmp)
peaksDt[!is.na(module_id_tmp), module_id:=moduleIds[module_id_tmp]]
peaksDt[is.na(module_id), module_id:=BG_NAME]
peaksDt[, module_id_tmp:=NULL]
moduleIds <- structure(peaksDt$module_id, names=peaksDt$rid)

setkey(dA, library_name)	
setkey(peaksDt, rid)

# DAR modules:
regionSetsOfInterest <- c(regionSetsOfInterest, sapply(split(peaksDt[sig==T,], by="module_id"), dtToGr, metaCols=c("rid")))
```

### 2. Region modules R7-R9 = differential in MUT, but not during differentiation 

= "de-novo DARs"

```{r atac_denovo_1}
# pick "differential && module=="BG"
tmp <- sapply(2:length(CONDITION_ORDER), function(condI) {
	dtToGr(peaksDt[get(CONDITION_ORDER[condI])==TRUE & module_id=="BG",], metaCols="rid")
}, simplify=FALSE)
names(tmp) <- paste_(CONDITION_ORDER[-1],"denovo")
regionSetsOfInterest[names(tmp)] <- tmp

peaksDt <- merge(peaksDt, dcast(rblapply(tmp, function(x) data.table(rid=x$rid)), rid~id), by="rid", all.x=TRUE)
forEach(names(tmp), function(x) {
	peaksDt[, (x):=!is.na(get(x))]
})
```

```{r atac_denovo_2}
moduleRegs <- unique(unlist(sapply(tmp, function(x) x$rid)))
KK <- 3

tmp <- versionedCache("atac_modules_denovo", instruction={		

	hmData <- prepHeatmapData(input_data[module_regs, module_samples], do_scale=T)

	# * calculate distance matrix and perform clustering:
	distMat <- dist(hmData, method=dist_meth)
	rc <- fastcluster::hclust(distMat, clust_meth)
	
	return(list(hm_data=hmData, row_clust=rc))
	
}, buildEnvir=list(
  input_data=d, module_regs=moduleRegs, module_samples=moduleSamples, dist_meth=ATAC_DIST_METH, clust_meth=ATAC_CLUST_METH
))
```
Assign each DAR to one of those "modules":

```{r atac_denovo_3}
peaksDt[, module_id_tmp:=paste0("tmp",cutree(tmp$row_clust, k=KK)[rid])]
setkey(peaksDt, rid)
```

Figure out an order of region modules that puts the modules in roughly temporal order:

```{r atac_denovo_4}
darOrderTmp <- rownames(tmp$hm_data)[tmp$row_clust$order]
pData <- prepHeatmapData(d[darOrderTmp,], do_scale=T)
pData <- melt(as.data.table(pData, keep.rownames="rid"), variable.name="library_name", measure.vars=colnames(pData))
pData[, module_id_tmp:=peaksDt[rid, module_id_tmp]]
pData <- merge(pData, dA, by="library_name")

tmp <- dtToDf(dcast(pData[, mean(value), by=.(day_fac, module_id_tmp)], module_id_tmp~day_fac, value.var="V1"))[,DAY_ORDER]

modOrderTmp <- (rownames(tmp)[seriation::get_order(seriation::seriate(tmp))])
round(tmp[modOrderTmp,])

denovoOrder <- unlist(lapply(split.data.frame(d[darOrderTmp,], f=peaksDt[darOrderTmp, factor(module_id_tmp, levels=modOrderTmp)]), function(mat) {
	rownames(mat)[seriation::get_order(seriation::seriate(mat))]
}), use.names=F)
denovoModOrder <- paste0("R",(K+1):(K+KK))
denovoIds <- structure(denovoModOrder, names=modOrderTmp)
peaksDt[!is.na(module_id_tmp), denovo_id:=denovoIds[module_id_tmp]]
peaksDt[is.na(denovo_id), denovo_id:=BG_NAME]
peaksDt[, module_id_tmp:=NULL]
denovoIds <- structure(peaksDt$denovo_id, names=peaksDt$rid)
```
Define to be used for all modules:

```{r atac_denovo_5}
setkey(dA, library_name)	
setkey(peaksDt, rid)

peaksDt[, all_mods_id := module_id]
peaksDt[denovo_id!="BG", all_mods_id := denovo_id]
allModsOrder <- names(COLOR_PALETTES$all_mods_id)

# DAR modules:
regionSetsOfInterest <- c(regionSetsOfInterest, sapply(split(peaksDt[sig==T & denovo_id!="BG",], by="denovo_id"), dtToGr, metaCols=c("rid")))
```	

Define groups of ROIs to be juxtaposed in plots:

```{r cmps}
cmpSelections <- list(
	all_mods_id = names(COLOR_PALETTES$all_mods_id),
	mut_effects = c(CONDITION_ORDER[-1],"BG"),
	mut_effects_dir = c(paste_(CONDITION_ORDER[-1],"up"),paste_(CONDITION_ORDER[-1],"down"))
)

cmpSummary <- rblapply(cmpSelections, function(x) {
  print(x)
	rblapply(regionSetsOfInterest[x], function(y) data.table(n = length(y)), "id")	
}, "cmp")
fwrite(cmpSummary, file=resultsDir("cmp_summary.csv"))
print(cmpSummary)
```

Save caches to be used in follow-up analyses:

```{r save_caches}
cmpSelections <- versionedCache(CACHE_ATAC_COMPARISONS, instruction={ cmp_selections }, buildEnvir=list(cmp_selections=cmpSelections))
regionSetsOfInterest <- versionedCache(CACHE_ATAC_REGION_SETS, instruction={ rois }, buildEnvir=list(rois=regionSetsOfInterest))
peaksDt <- versionedCache(CACHE_ATAC_PEAKS_ANNOTATED, instruction={ annot_peaks }, buildEnvir=list(annot_peaks=peaksDt))
regionOrder <- versionedCache(CACHE_ATAC_REGION_ORDER, instruction={ c(dar_order,denovo_order) }, buildEnvir=list(dar_order=darOrder, denovo_order=denovoOrder))
```

### Heatmaps

Visualize acessibility across all DARs in different stages and conditions:

```{r dar_heatmap, fig.width=12, fig.height=8}
selCols <- grep("_(up|down)", colnames(peaksDt), value=TRUE)

i <- c(darOrder,denovoOrder)
hm <- plotHMorder(d[i,], row_cols=c("all_mods_id", paste_(CONDITION_ORDER[-1],"up"),paste_(CONDITION_ORDER[-1],"down")), n=sprintf("Chromatin Modules & De-Novo, n = %d", length(i)), row_lbls=rep("",length(i)), do_scale=TRUE, rc=FALSE, gaps_row=peaksDt[i, .N, by=factor(all_mods_id, levels=allModsOrder)][,cumsum(N)])

pdf(figuresDir("figure_atac_heatmap.pdf"), width=12, height=8)
ComplexHeatmap::draw(hm)
dev.off()

ComplexHeatmap::draw(hm)

fwrite(data.table(region_id=i, module_id=peaksDt[i,all_mods_id], round(d[i,],3), peaksDt[i,c(paste_(CONDITION_ORDER[-1],"up"),paste_(CONDITION_ORDER[-1],"down")),with=FALSE]), file = figuresDir("figure_atac_heatmap_data.csv"))
```



### Euler diagrams

Visualize overlap between DARs:

```{r dar_venn}
print(colSums(peaksDt[,c("17q","17q1q","17q1qMYCN")]))
print(plot(eulerr::euler(peaksDt[,c("17q","17q1q","17q1qMYCN")]))	)

pdf(figuresDir("figure_atac_venn.pdf"), width=3, height=2)
print(plot(eulerr::euler(peaksDt[,c("17q","17q1q","17q1qMYCN")]))	)
dev.off()

fwrite(peaksDt[,.(rid, `17q`=`17q`*1, `17q1q`=`17q1q`*1, `17q1qMYCN`=`17q1qMYCN`*1)], file = figuresDir("figure_atac_venn_data.csv"))
```


### Line plots

Visualize mean accessbility in per cell line and stage in each module:

```{r module_line_plots}
i <- c(darOrder,denovoOrder)

pAgg <- prepHeatmapData(d[i,], do_scale=FALSE)
pAgg <- rblapply(split.data.frame(pAgg, f=peaksDt[i, factor(all_mods_id, levels=allModsOrder)]), function(x) { 
	y <- colMeans(x, na.rm=T) 
	data.table(library_name=names(y), value=y)
}, "all_mods_id")
pAgg <- merge(pAgg, dA, by="library_name")

p <- ggplot(pAgg[all_mods_id!=BG_NAME, ], aes(x=day, y=value, color=condition)) + stat_summary(geom="point", size=0.5, fun.data="mean_se") + stat_summary(geom="line", fun.data="mean_se") + facet_wrap(all_mods_id~., nrow=3) + defTheme() + scale_color_manual(values=COLOR_PALETTES$condition, guide="none") + xlab(NULL) + ylab("Mean normalized read count") + themeBoxed(panel.grid.major.y=element_line(size=0.25)) + scale_x_continuous(breaks=as.numeric(gsub("D","", DAY_ORDER)))


ggsave(filename = figuresDir("figure_atac_lines.pdf"), plot = p, width = 6, height = 3, dpi = 600, unit = "in")
fwrite(pAgg[all_mods_id!=BG_NAME, .(module_id = all_mods_id, library_name, condition, day, value)], file = figuresDir("figure_atac_lines_data.csv"))

print(p)
```

### Pie charts

Summarize distribution of DARs over modules:

```{r module_numbers}

pData <- peaksDt[, .(N=.N, N_str=prettyIntegers(.N)), by=.(module=all_mods_id, chr17q=`17q`,chr17q1q=`17q1q`,chr17q1qMYCN=`17q1qMYCN`)]
pData <- pData[order(module, !`chr17q`,!`chr17q1q`,!`chr17q1qMYCN`),]

print(pData)

cmp <- setdiff(cmpSelections$mut_effects_dir,BG_NAME)

pData <- melt(peaksDt, measure.vars=cmp)
pData[value==TRUE, val:=variable]
pData[value==FALSE, val:=BG_NAME]

pDataX <- pData[, .(perc_of_module=sum(value)/.N, n=sum(value), .N), by=.(all_mods_id, mut=variable)][, .(perc_of_module,all_mods_id, n, N, perc_of_mut=n/sum(n), pos=cumsum(n/sum(n))), by=mut]

pDataX[, mut_simple:=gsub("_(up|down)","",mut)]

nByMut <- pDataX[, .(N=sum(n)), by=.(mut)]
nByMut <- structure(nByMut$N, names=as.character(nByMut$mut))

pDataX[grepl("down",mut), n:= -n]

pDataX <- pDataX[all_mods_id!="BG",]
p1 <- ggplot(pDataX, aes(x=1, y=perc_of_mut, fill=all_mods_id)) + geom_bar(stat="identity") + theme_void() + scale_y_continuous(expand=c(0,0)) + scale_fill_manual(values=repl(COLOR_PALETTES$all_mods_id,BG_NAME,"grey",by_name=T), guide="none") + ylab(NULL) + xlab(NULL) +
  coord_polar("y", start=0) + facet_wrap(~mut, nrow=2, labeller=function(x) data.frame(mut=sprintf("%s\nn = %s", x$mut, prettyIntegers(nByMut[as.character(x$mut)]))))
p3 <- ggplot(pDataX, aes(x=all_mods_id, y=n)) + geom_bar(aes(fill=all_mods_id), stat="identity") + defTheme(noLegendTitle=F, topLegend=T, flipX=T) + geom_hline(yintercept=0)	+ themeBoxed(panel.grid.major.y=element_line()) + facet_wrap(~mut) + scale_fill_manual(values=repl(COLOR_PALETTES$all_mods_id,BG_NAME,"grey",by_name=T), guide="none") + ylab("Peaks affected (n)") + xlab(NULL)  + facet_wrap(~mut_simple, nrow=1)

ggsave(filename = figuresDir("figure_atac_pies.pdf"), plot = p1, width = 6, height = 3, dpi = 600, unit = "in")
ggsave(filename = figuresDir("figure_atac_bars.pdf"), plot = p3, width = 6, height = 2, dpi = 600, unit = "in")
fwrite(pDataX, file = figuresDir("figure_atac_pies_and_bars_data.csv"))


print(p1)
print(p3)
```


# Appendix

**Runtime**

`r time_diff(SETUP_TIME)`

**Session Info**

```{r}
sessionInfo()
```
